---
title: "Homework, week 9"
author: "Peter Ralph"
date: "7 March 2019 -- Advanced Biological Statistics"
---

# Community structure

We have a dataset of abundances of 40 microbial taxa,
measured at 100 times (by 16S transcript counts)
throughout a decomposition experiment.
We are interested in whether there are groups of taxa that change in abundance together,
perhaps because they depend on each others' metabolic byproducts,
or perhaps because they are all responding to some common underlying variables that we can't observe
(e.g., amount of sugar left in the substrate as it decomposes).

Here is the dataset: [taxon_counts.csv](taxon_counts.csv)

The first column has the sample number.
The second column (labeled "time") has the time that the dataset was recorded at.
The remaining columns have the transcript counts for each taxon.

```{r setup, include=FALSE}
library(rstan)
library(matrixStats)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## 1.

Use `optimizing( )` to find a maximum posterior estimate of the parameters
in the following Stan model, with $K=3$ communities.
In this model, `w[i][j]` contains the proportion sample `i` that is made up of community `j`,
and `(eta .* x)[i,j]` contains the mean abundance of taxon `i` in community `j`.
Note that in this model, a given taxon may belong to more than one community 
(although with different relative abundances).
(Also: this is the *same* model that we used in analyzing the transcription data in class.)

```{r nmf, cache=TRUE}
nmf_model <- stan_model(model_code="
    data {
      int N; // # samples
      int L; // # taxa
      int K; // # clusters
      int Z[N,L];
    }
    parameters {
      matrix<lower=0>[L,K] x;
      vector[L] y;
      simplex[K] w[N];
      vector<lower=0>[L] eta;
      vector<lower=0>[K] alpha;
      real<lower=0> d_alpha;
    }
    model {
      for (i in 1:N) {
          Z[i] ~ poisson(eta .* (x * w[i]));
          w[i] ~ dirichlet(d_alpha * alpha);
      }
      for (j in 1:K) 
          { x[,j] ~ normal(y ./ eta, 1); }
      y ~ normal(0, 20);
      alpha ~ normal(0, 1);
      d_alpha ~ exponential(0.2);
      eta ~ cauchy(0, 10);
    }")
```

## 2.

Do the relative contributions of the three communities change over time?
Make a plot showing 
estimated abundances of each cluster of taxa against time.

## 3.

How do the communities differ?
Make a plot showing the abundances of each taxon in the three communities.


```{r sim_data, cache=TRUE, include=FALSE}
nclusters <- 3
nspp <- 40
nsamples <- 100
pinf <- 0.45 # number of informative species
signal_mean <- 100
backgr_mean <- 5.0
noise_sd <- 5.0

# time-series based proportions per sample
tt <- sort(runif(nsamples))
# Simulate proportions per sample
ew <- do.call(cbind, lapply(1:nclusters, function (k) {
                dnorm(tt, mean=(k-1)/(nclusters-1), sd=0.3)
            } ))
ew <- sweep(ew, 1, rowSums(ew), "/")
w <- ew * rgamma(nsamples*nclusters, rate=1/10, shape=10)
w <- sweep(w, 1, rowSums(w), "/")
stopifnot(all(abs(rowSums(w) - 1) < 1e-15))
# matplot(tt, ew, type='l')
# matpoints(tt, w)


# Determine mean abundances by species
mu <- rgamma(nspp, rate=10/signal_mean, shape=10)
# copy mu into cluster-specific columns
type_mu <- do.call(cbind, list(mu)[rep(1,nclusters)])

# simulate how different mu is across clusters
# which cluster do species fall in?
sp_type <- sample(0:nclusters, nspp, replace=TRUE, prob=c(pinf, (1-pinf)*c(.5, .3, .2)))
for (j in 1:nclusters) {
    for (k in 1:nclusters) {
        if (j != k) {
            type_mu[sp_type == k, j] <- rexp(sum(sp_type == k)) * backgr_mean
        }
    }
}

# construct expected expression levels:
em <- matrix(0, nrow=nsamples, ncol=nspp)
for (ct in 1:nclusters) {
    em <- em + outer(w[,ct], type_mu[,ct])
}
m <- matrix(pmax(0, em + rnorm(length(em), noise_sd)), nrow=nrow(em))

# simulate counts
Z <- matrix(rpois(length(m), m), ncol=ncol(m))
```
```{r write_data, include=FALSE}
colnames(Z) <- apply(matrix(sample(letters, nspp * 4, replace=TRUE), ncol=4), 1, paste, collapse='')

write.csv(cbind(time=tt, Z), file="taxon_counts.csv")
```

```{r analyze, cache=TRUE, dependson=c("nmf", "sim_data"), include=FALSE}
ot <- system.time(nmf_optim <- optimizing(nmf_model,
                                    data=list(N=nsamples,
                                              L=nspp,
                                              K=nclusters,
                                              Z=Z)))
```


```{r results, include=FALSE}
results <- list(x=nmf_optim$par[grepl("^x", names(nmf_optim$par))],
                y=nmf_optim$par[grepl("^y", names(nmf_optim$par))],
                w=nmf_optim$par[grepl("^w", names(nmf_optim$par))],
                eta=nmf_optim$par[grepl("^eta", names(nmf_optim$par))],
                alpha=nmf_optim$par[grepl("^alpha", names(nmf_optim$par))],
                d_alpha=nmf_optim$par[grepl("^d_alpha", names(nmf_optim$par))])
dim(results$x) <- c(nspp, nclusters)
dim(results$w) <- c(nsamples, nclusters)

matplot(tt, results$w,
        xlab='time', ylab='proportion')
```
