---
title: "Heights"
author: "Peter Ralph"
date: "October 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's recreate Galton's classic analysis:
midparent height, adjusted for gender, is a good predictor of child height.
*(How good?)*

Link to [the data](../Datasets/galton/galton-all.tsv).
```{r read_galton}
galton <- read.table("../Datasets/galton/galton-all.tsv", header=TRUE)
head(galton)
```

## Look at the data

Here are the heights of children:
```{r look}
hist(galton$height, breaks=30, xlab='height', main='')
```

Now let's look at heights by sex:
```{r more_look}
layout((1:2))
hist(galton$height[galton$gender=="M"], breaks=20, xlab='height', main='', xlim=range(galton$height))
hist(galton$height[galton$gender=="F"], breaks=20, xlab='height', main='', xlim=range(galton$height))
```

Also, we should look at parent and child height relationships:
red is male, black is female; circles are fathers, squares are mothers.
```{r moremorelook}
plot(father ~ height, data=galton, col=gender, ylim=range(father, mother), ylab='parent heights', pch=20, xlab='child height', asp=1)
points(mother ~ height, data=galton, col=gender, pch=22)
```

Is there a correlation between parents' heights?
```{r parent_reg}
plot(jitter(father) ~ jitter(mother), data=galton, asp=1, col=adjustcolor("black", 0.5), pch=20)
```
Not obviously.

## Adjust by gender

To adjust by sex, we need to, separately for parents and children,

- compute the mean of male and female height
- subtract the difference from males

```{r adjust}
parent_means <- c('father' = mean(tapply(galton$father, galton$family, unique)),
                  'mother' = mean(tapply(galton$mother, galton$family, unique)))
child_means <- c('F' = mean(subset(galton, gender=="F")$height),
                 'M' = mean(subset(galton, gender=="M")$height))
galton$adj_father <- galton$father + diff(parent_means)
galton$adj_height <- galton$height - ifelse(galton$gender == "F", 0, abs(diff(child_means)))
```

Now the histograms line up:
```{r new_hist}
layout((1:2))
hist(galton$adj_height[galton$gender=="M"], breaks=20, xlab='height', main='', xlim=range(galton$adj_height))
hist(galton$adj_height[galton$gender=="F"], breaks=20, xlab='height', main='', xlim=range(galton$adj_height))
```

## Compute midparent height

The midparent height is the average of the *adjusted* father height and the mother height.
```{r add_midparent}
galton$midparent <- (galton$adj_father + galton$mother)/2
```

## Fit a model


```{r do_lm}
the_lm <- lm(adj_height ~ midparent, data=galton)
summary(the_lm)
```

Diagnostics: first, residuals versus fitted:
```{r rf}
plot(fitted(the_lm), resid(the_lm), xlab='fitted values',
     ylab='residuals')
abline(h=0, lty=3)
```

Look at residual distribution:
```{r qqplot}
qqnorm(resid(the_lm))
qqline(resid(the_lm))
```

## Look for residual signal

It doesn't look like we're missing anything obvious:
```{r big_lm}
summary(lm(adj_height ~ midparent + I(father - mother) + kids + gender, data=galton))
anova(lm(adj_height ~ midparent, data=galton),
      lm(adj_height ~ midparent + I(father - mother) + kids + gender, data=galton))

```



