
# Missing data: imputation

From Week 10 - NOT WORKING

## Imputation

. . .

is a fancy word for "guessing".

. . .

Often, if some data are missing, it's useful to guess what it probably should be

. . .

... for instance, by sampling from the posterior distribution
of your missing data given the rest of the data.

## For instance

Suppose that we're missing the *ages* of some of our samples.
Can we <strike>guess</strike> impute the missing ages?

## 

```{r impute_model, cache=TRUE}
impute_model <- stan_model(model_code="
data {
    int N;  // number of data points
    int Nm; // number of missing ages
    vector[N - Nm] notmissing_age;
    real minage;
    vector[N] exposure;
    int missings[Nm]; // indexes of missings
    int not_missings[N - Nm]; // the complement
    int counts[N];
    int genotype[N];
    int ngenotypes;
}
parameters {
    vector[ngenotypes] a; // intercepts
    real b; // slope for age
    vector[ngenotypes] c; // slopes for exposure
    real<lower=0> sigma; // SD on lognormal
    vector<lower=0>[N] mu; // mean of the poissons
    vector<lower=0>[Nm] missing_age;
}
model {
    vector[N] age;
    vector[N] y; // mean of the lognormals
    age[missings] = missing_age;
    age[not_missings] = notmissing_age;
    y = a[genotype] + b * age + c[genotype] .* exposure;
    mu ~ lognormal(y, sigma);
    counts ~ poisson(mu);
    a ~ normal(0, 100);
    b ~ normal(0, 10);
    c ~ normal(0, 20);
    sigma ~ normal(0, 10);
    missing_age ~ normal(0, 1);
}
")
```


##

```{r impfit, cache=TRUE, dependson="impute"}
impute_these <- sort(sample(length(data$counts), 40))

impute_data <- with(data,
      list(N=length(counts),
           Nm=length(impute_these),
           notmissing_age=(age[-impute_these] - mean(age))/sd(age),
           minage=(0 - mean(age))/sd(age),
           exposure=(exposure - mean(exposure))/sd(exposure),
           counts=counts,
           missings=impute_these,
           not_missings=seq_along(counts)[-impute_these],
           genotype=genotype,
           ngenotypes=length(unique(genotype))))

fit4 <- sampling(impute_model,
                 data=impute_data,
                 control=list(max_treedepth=12),
                 iter=100, chains=1)
```

##

```{r plot_imp, fig.width=3*fig.dim, fig.height=2*fig.dim, echo=FALSE}
imps <- mean(data$age) + sd(data$age) * rstan::extract(fit4, pars="missing_age")$missing_age
ord <- rank(colMeans(imps))
plot(ord, colMeans(imps), ylim=c(0,150), xlab='', ylab='age')
segments(x0=ord,
         y0=colMins(imps),
         y1=colMaxs(imps), lty=3, lwd=0.5)
segments(x0=ord,
         y0=colQuantiles(imps, probs=0.025),
         y1=colQuantiles(imps, probs=0.975), col='red')
segments(x0=ord,
         y0=colQuantiles(imps, probs=0.25),
         y1=colQuantiles(imps, probs=0.75), lwd=2)
points(x=ord, y=data$age[impute_these], cex=2, pch=20)
legend("topleft",
       lty=c(3,1,1,NA), pch=c(NA,NA,NA,20), col=c("black", "red", "black", "black"),
       pt.cex=c(NA,NA,NA,2), lwd=c(3,1,1,NA),
       legend=c("range", "95% CI", "50% CI", "true value"))
```


