---
title: "The Beta distribution"
author: "Peter Ralph"
date: "11 November -- Advanced Biological Statistics"
---


```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(matrixStats)
```



# Stochastic Minute

## The Beta Distribution

If $$P \sim \text{Beta}(a,b)$$
then $P$ has *probability density*
$$p(\theta) = \frac{ \theta^{a-1} (1 - \theta)^{b-1} }{ B(a,b) } . $$

- Takes values between 0 and 1.

- If $U_{(1)} < U_{(2)} < \cdots < U_{(n)}$ are sorted, independent $\text{Unif}[0,1]$
  then $U_{(k)} \sim \text{Beta}(k, n-k+1)$.

- Mean: $a/(a+b)$.

- Larger $a+b$ is more tightly concentrated (like $1/\sqrt{a+b}$)

##

```{r beta_stuff, echo=FALSE, fig.width=2.5*fig.dim, fig.height=1.5*fig.dim}
shadecurve <- function (pf, xlim, plot=TRUE, xlab='', ylab='', main='', yaxt='n',
                        border="black", col=adjustcolor(border, 0.25), ...) {
    x <- seq(xlim[1], xlim[2], length.out=401)
    mids <- x[-1] - diff(x)/2
    df <- diff(pf(x, ...))
    if (plot) { plot(0, type='n', xlim=range(x), ylim=range(df),
                     main=main, xlab=xlab, ylab=ylab, yaxt=yaxt) }
    polygon(c(mids, x[length(x)], x[1]), c(df, 0, 0), col=col, border=border) 
    return(invisible(list(x=x, y=df)))
}

a <- 20
b <- 40
par(mar=c(3,1,3,0)+.1)
xy <- shadecurve(pbeta, c(0, 1), shape1=a, shape2=b, border="blue", xlab='', ylab='')
title(main="Beta(a, b)")
abline(v=a/(a+b), lty=3, lwd=2)
text(x=a/(a+b)+0.05, y=max(xy$y)*0.8,
     labels=expression(mu == frac(a,a+b)), pos=4)
qy <- xy$y[findInterval(qbeta(0.05, a, b), xy$x)]
lines(x=xy$x[diff(xy$y > qy) != 0],
      y=rep(xy$y[diff(xy$y > qy) > 0], 2), lwd=2)
text(x=xy$x[diff(xy$y > qy) < 0],
     y=xy$y[diff(xy$y > qy) > 0],
     labels=expression(sqrt(frac(mu * (1-mu), a + b + 1))), pos=4)
```



## Beta-Binomial Bayesian analysis

If
$$\begin{aligned}
P &\sim \text{Beta}(a,b) \\
Z &\sim \text{Binom}(n,P) ,
\end{aligned}$$
then by Bayes' rule:
$$\begin{aligned}
 \P\{ P = \theta \given Z = z\}
 &=
 \frac{\P\{Z = z \given P = \theta \} \P\{P = \theta\}}{\P\{Z = z\}} \\
 &= 
 \frac{
 \binom{n}{z}\theta^z (1-\theta)^{n-z}
 \times
 \frac{\theta^{a-1}(1-\theta)^{b-1}}{B(a,b)}
 }{
 \text{(something)}
 } \\
 &=
 \text{(something else)} \times \theta^{a + z - 1} (1-\theta)^{b + n - z - 1} .
\end{aligned}$$

-------------

"Miraculously"\
*(the Beta is the conjugate prior for the Binomial)*,
$$\begin{aligned}
(P \given Z = z) \sim \text{Beta}(a+z, b+n-z) .
\end{aligned}$$

## Exercise

*Simulate:*

1. One thousand "random coins" whose probabilities are drawn from a $\Beta(5,5)$ distribution.
    (`rbeta()`)  Make a histogram of these probabilities.

2. Flip each coin ten times and record the number of heads. (`rbinom()`)

3. Make a histogram of the *probabilities* of those coins
    that got exactly 3 heads, and compare to the first histogram.


