---
title: "The $t$ distribution"
author: "Peter Ralph"
date: "1 October 2020 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
```


# Stochastic minute: the $t$ distribution

## The $t$ statistic

The $t$ statistic computed from a collection of $n$ numbers
is the sample mean divided by the estimated standard error of the mean,
which is the sample SD divided by $\sqrt{n}$.

. . .

If $x_1, \ldots, x_n$ are numbers, then
$$\begin{aligned}
    \text{(sample mean)} \qquad \bar x &= \frac{1}{n}\sum_{i=1}^n x_i \\
    \text{(sample SD)} \qquad s &= \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar x)^2} 
\end{aligned}$$
so
$$\begin{equation}
    t(x) = \frac{\bar x}{s / \sqrt{n}} .
\end{equation}$$

## Consistency check

```{r t_check}
n <- 20
x <- rnorm(n)
c(t.test(x)$statistic, 
  mean(x) / (sd(x) / sqrt(n)))
```

## The $t$ approximation

**Fact:** 
If $X_1, \ldots, X_n$ are independent random samples from a distribution with mean $\mu$,
then
$$\begin{equation}
    t(X - \mu) = \frac{\bar x - \mu}{s/\sqrt{n}} \approx \StudentsT(n-2) ,
\end{equation}$$
as long as $n$ is not too small and the distribution isn't too wierd.

## A demonstration

Let's check this, by doing:

> find the sample $t$ score of 100 random draws from some distribution

lots of times, and looking at the distribution of those $t$ scores.

. . .

Claim: no matter${}^*$ the distribution we sample from,
the *sampling distribution* of the $t$ statistics should look close to
the $t$ distribution.

## One sample

```{r t_one_smaple}
n <- 20
x <- 2 * runif(n) - 1
hist(x, xlab='value', col=grey(0.5),
     main=sprintf("t=%f", mean(x)*sqrt(n)/sd(x)))
abline(v=0, lwd=2, lty=3)
abline(v=mean(x), col='red', lwd=2)
```

## More samples

```{r t_more_samples, echo=FALSE, fig.height=2*fig.dim, fig.width=2.5*fig.dim}
par(mar=c(4,3,1,1))
layout(matrix(1:20, nrow=4))
for (k in 1:20) {
    x <- 2 * runif(n) - 1
    hist(x, xlab='value', col=grey(0.5),
         main=sprintf("t=%f", mean(x)*sqrt(n)/sd(x)))
    abline(v=0, lwd=2, lty=3)
    abline(v=mean(x), col='red', lwd=3)
}
```

## Distribution of 1,000 sample $t$ scores

```{r t_sampling_dist}
xm <- replicate(1000, {
            x <- 2 * runif(n) - 1;
            mean(x) * sqrt(n) / sd(x) })
xh <- hist(xm, breaks=40, main=sprintf('t of %d samples', n), col='red')
```

## Distribution of 1,000 sample $t$ scores

```{r t_smpling_dist2}
plot(xh, main=sprintf('t of %d samples', n), col='red')
xx <- xh$breaks
polygon(c(xx[-1] - diff(xx)/2, xx[1]),
        c(length(xm)* diff(pt(xx, df=(n-1))), 0),
        col=adjustcolor("blue", 0.4))
```

## Exercise:

Do this again (use my code) except using
```
x <- rexp(n) - 1
```
instead of `2 * runif(n) - 1`.

