---
title: "Categorical data and permutation tests"
author: "Peter Ralph"
date: "21 January 2020 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(matrixStats)
library(tidyverse)
options(digits=2)
```

# Hair and eye color

## Hair and Eye color

```{r hair_eye_data}
data(HairEyeColor)
```
```
HairEyeColor             package:datasets              R Documentation

Hair and Eye Color of Statistics Students

Description:

     Distribution of hair and eye color and sex in 592 statistics
     students.

Usage:

     HairEyeColor
     
Format:

     A 3-dimensional array resulting from cross-tabulating 592
     observations on 3 variables.  The variables and their levels are
     as follows:

       No  Name  Levels                    
        1  Hair  Black, Brown, Red, Blond  
        2  Eye   Brown, Blue, Hazel, Green 
        3  Sex   Male, Female              
      
Details:

     The Hair x Eye table comes from a survey of students at the
     University of Delaware reported by Snee (1974).  The split by
     ‘Sex’ was added by Friendly (1992a) for didactic purposes.

     This data set is useful for illustrating various techniques for
     the analysis of contingency tables, such as the standard
     chi-squared test or, more generally, log-linear modelling, and
     graphical methods such as mosaic plots, sieve diagrams or
     association plots.

Source:

     <URL:
     http://euclid.psych.yorku.ca/ftp/sas/vcd/catdata/haireye.sas>

     Snee (1974) gives the two-way table aggregated over ‘Sex’.  The
     ‘Sex’ split of the ‘Brown hair, Brown eye’ cell was changed to
     agree with that used by Friendly (2000).

References:

     Snee, R. D. (1974).  Graphical display of two-way contingency
     tables.  _The American Statistician_, *28*, 9-12.  doi:
     10.2307/2683520 (URL: http://doi.org/10.2307/2683520).

```

----------------

```{r show_he, echo=FALSE}
HairEyeColor
```

---------------

```{r melt_he}
haireye <- as.data.frame(HairEyeColor)
names(haireye) <- tolower(names(haireye))
names(haireye)[names(haireye) == "freq"] <- "number"
haireye
```

## 

*Questions:* 

1. Are hair and eye color independent in this sample?
2. Do hair and eye color proportions differ by sex?



# Method 2: Permutation

## recall the $p$-value

. . .

... is the probability of seeing something at least as extreme as what we saw in the data,
if the null hypothesis (model) is true.

. . .

A permutation test *estimates* the "probability ... under the null hypothesis" part.

. . .

- "the null hypothesis" is that *hair and eye color are independent*.
- we can use... the chi-squared statistic for "at least as extreme"?

##

First, "individualize" the data:
```{r long_data}
long_haireye <- haireye[rep(1:nrow(haireye), haireye$number), 
                        c("hair", "eye", "sex")]
stopifnot(nrow(long_haireye) == sum(haireye$number))
long_haireye
```

##

Compute the chi-squared statistic with `chisq.test( )`:
```{r do_chisq2, warning=FALSE}
he_tab <- table(long_haireye[long_haireye$sex == 'Female', 1:2])
csq <- chisq.test(he_tab)
str(csq)
```

##  Are hair and eye independent, given sex?

```{r test_stat, warning=FALSE}
true_val <- 0
for (s in levels(long_haireye$sex)) {
    true_val <- (true_val + 
        with(subset(long_haireye, sex==s), 
                chisq.test(table(hair, eye))$statistic))
}
```

## Permutations:

```{r perm_test, warning=FALSE}
nperm <- 1000
chisq_perm <- rep(0, nperm)
for (k in 1:nperm) {
    for (s in levels(long_haireye$sex)) {
        fake <- subset(long_haireye, sex == s)
        fake$eye <- sample(fake$eye)
        chisq_perm[k] <- chisq_perm[k] + chisq.test(table(fake$eye, fake$hair))$statistic
    }
}
```

## Result:

```{r plot_perms, echo=FALSE}
hist(chisq_perm, breaks=40, xlab=expression(chi^2), xlim=c(0, 1.2 * true_val), main='bootstrap distribution of chi-squared value')
abline(v=true_val, col='red', lwd=2)
legend("topright", lty=1, col=2, lwd=2, legend="true value")
```

## Conclusion?

(What did we actually test?)


## Your turn:

Use a permutation test to assess
whether the relation between hair and eye color differs by sex.

To do this, we'll:

1. Compute the sum of squared differences
   between each of the 16 proportions (4 eye x 4 hair colors).
   *(test statistic)*
   
2. Compute the same thing, after shuffling up the `sex` column, a bunch of times.

3. Compare the observed value to the distribution under permutations from (2).


