---
title: "Categorical data and the chi-squared test"
author: "Peter Ralph"
date: "21 January 2020 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(tidyverse)
library(matrixStats)
options(digits=2)
```


# Categorical data

## Hair and Eye color

```{r hair_eye_data}
data(HairEyeColor)
```
```
HairEyeColor             package:datasets              R Documentation

Hair and Eye Color of Statistics Students

Description:

     Distribution of hair and eye color and sex in 592 statistics
     students.

Usage:

     HairEyeColor
     
Format:

     A 3-dimensional array resulting from cross-tabulating 592
     observations on 3 variables.  The variables and their levels are
     as follows:

       No  Name  Levels                    
        1  Hair  Black, Brown, Red, Blond  
        2  Eye   Brown, Blue, Hazel, Green 
        3  Sex   Male, Female              
      
Details:

     The Hair x Eye table comes from a survey of students at the
     University of Delaware reported by Snee (1974).  The split by
     ‘Sex’ was added by Friendly (1992a) for didactic purposes.

     This data set is useful for illustrating various techniques for
     the analysis of contingency tables, such as the standard
     chi-squared test or, more generally, log-linear modelling, and
     graphical methods such as mosaic plots, sieve diagrams or
     association plots.

Source:

     <URL:
     http://euclid.psych.yorku.ca/ftp/sas/vcd/catdata/haireye.sas>

     Snee (1974) gives the two-way table aggregated over ‘Sex’.  The
     ‘Sex’ split of the ‘Brown hair, Brown eye’ cell was changed to
     agree with that used by Friendly (2000).

References:

     Snee, R. D. (1974).  Graphical display of two-way contingency
     tables.  _The American Statistician_, *28*, 9-12.  doi:
     10.2307/2683520 (URL: http://doi.org/10.2307/2683520).

```

----------------

```{r show_he, echo=FALSE}
HairEyeColor
```

---------------

```{r melt_he}
haireye <- as.data.frame(HairEyeColor)
names(haireye) <- tolower(names(haireye))
names(haireye)[names(haireye) == "freq"] <- "number"
haireye
```

## 

*Questions:* 

1. Are hair and eye color independent in this sample?
2. Do hair and eye color proportions differ by sex?


## Independence and multiplicativity

If hair and eye color are *independent*,
then probabilities of combinations are *multiplicative*:

$$\begin{aligned}
    &\P\{\text{black hair and blue eyes}\} \\
    &\qquad =
        \P\{\text{black hair}\} \times \P\{\text{blue eyes}\given\text{black hair}\} \\
\end{aligned}$$

. . .

which if independent is
$$\begin{aligned}
    &\hphantom{\P\{\text{black hair and blue eyes}\}} \\
    &\qquad =
        \P\{\text{black hair}\} \times \P\{\text{blue eyes}\}
\end{aligned}$$


## Multiplicativity

A model of *independence* will have a *multiplicative* form:
$$
    p_{ab} = p_a \times p_b .
$$



# The chi-squared statistic

##

Let's start by looking at *just* hair and eye color,
summing over sex:
```{r nosex}
(haireye_2d <- HairEyeColor[,,"Male"] + HairEyeColor[,,"Female"])
```

## Some questions

In this dataset...

1. What proportion have blonde hair?
2. What proportion have blue eyes?
3. If hair and eye color assort independently,
   what proportion do you expect to have both blonde hair and blue eyes?
   How many people would this be?
4. How many *actually* have both?
   Is this difference surprising?
5. Do the same for black hair and green eyes.

## "Expected" counts

Let
$$\begin{aligned}
    n_{ij} &= (\text{observed}_{ij}) \\
        &=(\text{observed number with hair $i$ and eye $j$}) \\
    E_{ij} &= (\text{expected}_{ij}) \\
          &=(\text{total number})
                \times(\text{proportion with hair $i$}) \\
          &\qquad  \times (\text{proportion with eye $j$}) \\
          &= n
            \times \left(\frac{n_{i\cdot}}{n}\right)
            \times \left(\frac{n_{\cdot j}}{n}\right) .
\end{aligned}$$

Here $n_{i \cdot}$ 
and $n_{\cdot j}$ are the *row* and *column sums*.

## 

We want to quantify how different the *observed* and *expected* are,
inversely weighted by their *noisiness*:
$$\begin{aligned}
    \sum_{ij} \left( \frac{ (\text{observed})_{ij} - (\text{expected})_{ij} }{ \SE[\text{observed}_{ij}] } \right)^2 
\end{aligned}$$

. . .

So, what is $\SE[\text{observed}_{ij}]$?

## What is $\SE[\text{observed}_{ij}]$?

Under the model of independence,
$$\begin{aligned}
    n_{ij} &\sim \Binom(n, p_i q_j) , \\
    \text{where}\quad
    p_i &= (\text{prob of hair color $i$}) \\
    q_j &= (\text{prob of eye color $j$}) .
\end{aligned}$$

. . .

So,
$$\begin{aligned}
    \sd[n_{ij}] = \sqrt{ n p_i q_j (1 - p_i q_j) } ,
\end{aligned}$$

. . .

... and so how about this
$$\begin{aligned}
    \SE[n_{ij}] 
        &\approx \sqrt{ n p_i q_j } \\
        &= \sqrt{(\text{expected}_{ij})} \qquad \ldots?
\end{aligned}$$

## The chi-squared statistic

$$\begin{aligned}
\chi^2 &=
    \sum_{ij} \frac{ \left((\text{observed})_{ij} - (\text{expected})_{ij} \right)^2 }{ (\text{expected})_{ij} } .
\end{aligned}$$

i.e., "observed minus expected squared, divided by expected".

. . . 

This gives us a number. What does it mean?


# Chi-squared test for independence

## A chi-squared test

```{r chisq_test}
chisq.test(haireye_2d)
```

. . .

Um, ok? Hair and eye color are not independent?

## More context

Let's actually look at "observed minus expected":

```{r obsmexp, echo=1:2}
haireye_exp <- 0 * haireye_2d
haireye_exp[] <- ( rowSums(haireye_2d)[row(haireye_exp)]
                  * colSums(haireye_2d)[col(haireye_exp)]
                  / sum(haireye_2d) )
haireye_exp
```

##

:::: {.columns}
::::::::: {.column width="50%"}

Observed minus expected:
```{r ome, echo=FALSE}
(haireye_2d - haireye_exp)
```

::::
::::::::: {.column width="50%"}

Normalized by $\sqrt{\text{expected}}$:
```{r omes, echo=FALSE}
((haireye_2d - haireye_exp) / sqrt(haireye_exp))
```

::::
:::::::::

## Conclusions?


## What about by sex?

Compute the chi-squared statistic with `chisq.test( )`:
```{r do_chisq}
chisq.test(HairEyeColor[,,"Female"])
```

