---
title: "The logistic model"
author: "Peter Ralph"
date: "24 November 2020 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(tidyverse)
library(rstan)
library(matrixStats)
options(mc.cores = parallel::detectCores())
```

```{r last_time, cache=TRUE, include=FALSE}
batting <- read.csv("data/BattingAverage.csv", header=TRUE, stringsAsFactors=TRUE)
pos_model_block <- "
data {
    int N;   // number of players
    int hits[N];
    int at_bats[N];
    int npos; // number of positions
    int position[N];
}
parameters {
    real<lower=0, upper=1> theta[N];
    real<lower=0, upper=1> mu[npos];
    real<lower=0> kappa[npos];
}
model {
    real alpha;
    real beta;
    hits ~ binomial(at_bats, theta);
    for (i in 1:N) {
        alpha = mu[position[i]] * kappa[position[i]];
        beta = (1 - mu[position[i]]) * kappa[position[i]];
        theta[i] ~ beta(alpha, beta);
    }
    mu ~ beta(1,1);
    kappa ~ gamma(0.1,0.1);
} "
pos_model <- stan_model(model_code=pos_model_block)
pos_fit <- sampling(pos_model, chains=3, iter=1000,
                    data=list(N=nrow(batting),
                              hits=batting$Hits,
                              at_bats=batting$AtBats,
                              npos=nlevels(batting$PriPos),
                              position=as.numeric(batting$PriPos)))
param_samples <- extract(pos_fit)
batting$post_mean <- colMeans(param_samples$theta)
```


# The logistic model, a.k.a. "logistic regression"

## Motivation

There's a good bit [more information available](BattingAveragePlus.csv) about the baseball players.
```{r more_batting, cache=TRUE}
batplus <- read.csv("data/BattingAveragePlus.csv",
                    header=TRUE, stringsAsFactors=TRUE)
head(batplus)
```

. . .

Is batting average predicted by weight and height?

-------------

1. Each player has their own batting avg, $\theta_i$.

2. Batting averages tend to be different for different positions.

3. After accounting for position ($p_i$),
   $\theta$ varies with height ($h_i$) and/or weight ($w_i$).

# Relating values to probabilities

## The logistic function

$$\begin{aligned}
   f(x) = \frac{1}{1 + e^{-x}} . 
\end{aligned}$$

```{r logistic_intro, echo=FALSE, fig.width=2*fig.dim, fig.height=1.5*fig.dim}
curve(1/(1+exp(-x)), from=-5, to=5, main="logistic function")
```

## ... is the inverse of the logit function


$$\begin{aligned}
   f^{-1}(x) = \log\left(\frac{x}{1-x}\right) . 
\end{aligned}$$

```{r logit, echo=FALSE, fig.width=2*fig.dim, fig.height=1.5*fig.dim}
curve(log(x/(1-x)), from=0.005, to=0.995, main="logit function")
```

# Back to Stan

##

::::::::: {.columns}
::: {.column width=50%}

1. Each player has their own batting avg, $\theta_i$.

2. Batting averages tend to be different for different positions.

3. After accounting for position ($p_i$),
   $\theta$ varies with height ($h_i$) and/or weight ($w_i$).

```{r logistic, echo=FALSE}
curve(1/(1+exp(-x)), from=-5, to=5, main="logistic function")
```

:::::::::
::: {.column width=50%}

$$\begin{aligned}
    Z_i &\sim \Binom(N_i, \theta_i) \\
    \theta_i &\sim \Beta(\mu_i \kappa_{p_i}, (1-\mu_i) \kappa_{p_i}) \\
    \mu_i &= \logistic(\gamma_{p_i} + \alpha_h h_i + \alpha_w w_i) \\
    \kappa_p &\sim \Gam(0.1, 0.1) \\
    \gamma_{p} &\sim \Normal(0, 5) \\
    \alpha_h &\sim \Normal(0, 5) \\
    \alpha_w &\sim \Normal(0, 5) 
\end{aligned}$$

:::
:::::::::


## Questions:

1. How much do we expect $\gamma$ to differ for pitchers versus catchers?

2. What would a value of $\alpha_h = 1$ mean?

3. What would be a range of plausible values for $\alpha_h$ and $\alpha_w$?

4. Are the priors sensible?

. . .

*Note:* we'll **center and scale** weight and height,
so that $h$ and $w$ will be in units of *standard deviations from the mean*.


## Stan

::::::::: {.columns}
::: {.column width=50%}

```{r stan_logistic, cache=TRUE}
logistic_model <- "
data {
    int N;   // number of players
    int hits[N];
    int at_bats[N];
    int npos; // number of positions
    int position[N];
    vector[N] height;
    vector[N] weight;
}
parameters {
    real<lower=0, upper=1> theta[N];
    vector[npos] gamma;
    real alpha_w;
    real alpha_h;
    vector<lower=0>[npos] kappa;
}
model {
    vector[N] mu;
    mu = inv_logit(gamma[position] + alpha_h * height + alpha_w * weight);
    theta ~ beta(mu .* kappa[position], (1-mu) .* kappa[position]);
    hits ~ binomial(at_bats, theta);
    kappa ~ gamma(0.1, 0.1);
    gamma ~ normal(0, 5);
    alpha_h ~ normal(0, 5);
    alpha_w ~ normal(0, 5);
} "
```

:::::::::
::: {.column width=50%}

$$\begin{aligned}
    Z_i &\sim \Binom(N_i, \theta_i) \\
    \theta_i &\sim \Beta(\mu_i \kappa_{p_i}, (1-\mu_i) \kappa_{p_i}) \\
    \mu_i &= \logistic(\gamma_{p_i} + \alpha_h h_i + \alpha_w w_i) \\
    \kappa_p &\sim \Gam(0.1, 0.1) \\
    \gamma_{p} &\sim \Normal(0, 5) \\
    \alpha_h &\sim \Normal(0, 5) \\
    \alpha_w &\sim \Normal(0, 5) 
\end{aligned}$$

:::
:::::::::

----------

## New features

::::::::: {.columns}
::: {.column width=50%}

```r
logistic_model <- "
data {
    int N;   // number of players
    int hits[N];
    int at_bats[N];
    int npos; // number of positions
    int position[N];
    vector[N] height;
    vector[N] weight;
}
parameters {
    real<lower=0, upper=1> theta[N];
    vector[npos] gamma;
    real alpha_w;
    real alpha_h;
    vector<lower=0>[npos] kappa;
}
model {
    vector[N] mu;
    mu = inv_logit(gamma[position] + alpha_h * height + alpha_w * weight);
    theta ~ beta(mu .* kappa[position], (1-mu) .* kappa[position]);
    hits ~ binomial(at_bats, theta);
    kappa ~ gamma(0.1, 0.1);
    gamma ~ normal(0, 5);
    alpha_h ~ normal(0, 5);
    alpha_w ~ normal(0, 5);
} "
```

:::::::::
::: {.column width=50%}

1. `vector[N] height;` instead of `real height[N];`: to allow `alpha_h * height`.

2. `gamma[position]` for vectorized indexing

3. `x .* y` for component-wise multiplication

:::
:::::::::


## 

```
data {
    int N;   // number of players
    int hits[N];
    int at_bats[N];
    int npos; // number of positions
    int position[N];
    vector[N] height;
    vector[N] weight;
}
parameters {
    real<lower=0, upper=1> theta[N];
    vector[npos] gamma;
    real alpha_w;
    real alpha_h;
    vector<lower=0>[npos] kappa;
}
model {
    vector[N] mu;
    mu = inv_logit(gamma[position] + alpha_h * height + alpha_w * weight);
    theta ~ beta(mu .* kappa[position], (1-mu) .* kappa[position]);
    hits ~ binomial(at_bats, theta);
    kappa ~ gamma(0.1, 0.1);
    gamma ~ normal(0, 5);
    alpha_h ~ normal(0, 5);
    alpha_w ~ normal(0, 5);
} 
```



-------------------


```{r fit_logistic_model, cache=TRUE, dependson=c("stan_logistic", "more_batting")}
scaled_height <- (batplus$height - mean(batplus$height))/sd(batplus$height)
scaled_weight <- (batplus$weight - mean(batplus$weight))/sd(batplus$weight)
system.time(logistic_fit <- stan(model_code=logistic_model, chains=3, iter=1000,
                               control=list(adapt_delta=0.9, max_treedepth=12),
                               data=list(N=nrow(batplus),
                                      hits=batplus$Hits,
                                      at_bats=batplus$AtBats,
                                      npos=nlevels(batplus$PriPos),
                                      position=as.numeric(batplus$PriPos),
                                      height=scaled_height,
                                      weight=scaled_weight) ))
```

------------------

```{r results}
logistic_summary <- rstan::summary(logistic_fit)$summary
print(logistic_summary)
```

## Pitchers still don't bat well

```{r logistic_gamma}
stan_plot(logistic_fit, pars="gamma")
```

----------------

Recall the *mean* of $\theta_i$ is
$$\begin{aligned}
\mu_i = \frac{1}{1 + \exp(-\gamma_{p_i} - \alpha_h h_i - \alpha_w w_i)} .
\end{aligned}$$

```{r logistic_mu}
gamma_samples <- rstan::extract(logistic_fit)$gamma
colnames(gamma_samples) <- levels(batplus$PriPos)
mu_samples <- 1/(1+exp(-gamma_samples))
boxplot(mu_samples, ylab="posterior distrn, mean batting avg", las=3)
```


## No effect of height or weight

```{r logistic_alpha}
stan_plot(logistic_fit, pars=c("alpha_w", "alpha_h")) + geom_vline(xintercept=0)
```

## Same $\theta$ as before?

```{r log_theta, fig.width=1.5*fig.dim, fig.height=1.5*fig.dim}
new_post_theta <- colMeans(rstan::extract(logistic_fit)$theta)
old_post_theta <- batting$post_mean[match(batplus$Player, batting$Player)]
plot(old_post_theta, new_post_theta, xlab="theta, model 1", ylab="theta, model 2")
abline(0,1, col='red')
```


## Conclusions

1. We don't have good evidence that batting average varies substantially
   with height or weight.

2. Changing the model from a Beta prior on class means
   to a logistic transform of a Normal prior
   did not substantially affect the results.
