---
title: "Random effects, and mixed models"
author: "Peter Ralph"
date: "27 October -- Advanced Biological Statistics"
---


```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(tidyverse)
library(matrixStats)
library(lme4)
```


```{r pumpkins, include=FALSE}
pumpkins <- read.table("data/pumpkins.tsv", header=TRUE)
pumpkins$plot <- factor(pumpkins$plot)
head(pumpkins)
```

# Random effects

## An example: urchins eat algae

From Logan:

> To investigate density-dependent grazing effects of sea urchin Andrew and Underwood
> (1993) on filamentous algae measured the percentage of filamentous algae within five
> quadrats randomly positioned within each of four random patches of reef that were in turn
> nested within four sea urchin density treatments (no urchins, 33% of natural density, 66%
> natural density and 100% natural density). The sea urchin density treatment was considered
> a fixed factor and patch within density treatment as well as the individual quadrats were
> treated as random factors.

## An example: urchins eat algae

```{r andrewdata}
andrew_data <- read.table('../Datasets/Logan_data/andrew.tsv', header=T, sep='\t')
head(andrew_data)
```

There are four variables: `TREAT`, `PATCH`, `QUAD` and `ALGAE`

Main effect factor: `TREAT`

. . .

Both `QUAD` and `PATCH` are factors:
```{r factorit}
andrew_data$QUAD <- factor(andrew_data$QUAD)
andrew_data$PATCH <- factor(andrew_data$PATCH)
andrew_data$TREAT <- factor(andrew_data$TREAT, levels=c("0%", "33%", "66%", "100%"))
```



## Experimental design

```{r tableit}
with(andrew_data, table(PATCH, QUAD, TREAT))
```

## Response distribution

```{r boxit, fig.width=2.5*fig.dim, fig.height=1.5*fig.dim}
plot(ALGAE ~ TREAT, data=andrew_data)
points(ALGAE ~ jitter(as.numeric(TREAT)), data=andrew_data, pch=20, col=1+as.numeric(PATCH)%%4)
```

##

Why is this wrong?

```{r lme_urchins}
summary(lm(ALGAE ~ TREAT, data=andrew_data))
```

##

What we really want:
$$
\text{(algae)} = \text{(mean for treatment)} + \text{(mean offset for patch)} + \text{("noise")} .
$$

. . .

We *could* do:
```
ALGAE ~ TREAT + PATCH
```
... but do we care about all those patch means?

##

```{r do_biglm}
summary(lm(ALGAE ~ TREAT + PATCH, data=andrew_data))
```

## Random effects

Small modification:
$$
\text{(algae)} = \text{(mean for treatment)} + \text{(random offset for patch)} + \text{("noise")} .
$$

. . .

We add a *random intercept*:
```
ALGAE ~ TREAT + (1|PATCH)
```

## 

```{r do_lmer}
library(lme4)
alglm <- lmer(ALGAE ~ TREAT + (1|PATCH), data=andrew_data)
summary(alglm)
```

##

```{r compare_lmer}
anova(
      lmer(ALGAE ~ TREAT + (1|PATCH), data=andrew_data),
      lm(ALGAE ~ TREAT, data=andrew_data))
```

##

```{r compare_lmer2}
anova(
      lmer(ALGAE ~ TREAT + (1|PATCH), data=andrew_data),
      lmer(ALGAE ~ (1|PATCH), data=andrew_data))
```


## What *are* the random effects?

```{r ranef}
ranef(alglm)
```

##

```{r plot_ranef}
rfs <- ranef(alglm)$PATCH
ses <- rfs[,1] + outer(sqrt(as.vector(attr(rfs, "postVar"))), c(-2, 2), "*")
plot(rfs[,1], 1:nrow(rfs), xlab='patch mean', xlim=range(ses), ylab='')
segments(x0=ses[,1], x1=ses[,2], y0=1:nrow(rfs))
abline(v=0, col='red')
```

# Notes on mixed models

## The math is a lot harder.

For a simple linear model fit by `lm( )` (with *fixed effects*),
the *log-likelihood function* is just the sum of the squared residuals.

. . .

But with a *mixed model*, the likelihood *averages* over the values of the random effects,
which makes everything more difficult.

## You sometimes have to worry about *convergence*.

Since the math is harder,
mixed-model-fitting functions like `lmer( )` have to use various sorts of *numerical optimization* methods
to find the best-fitting parameters.

. . .

Sometimes, these may fail.

. . .

Notably, many use the [`REML`](https://en.wikipedia.org/wiki/Restricted_maximum_likelihood) approximation:
```
Usage:

     lmer(formula, data = NULL, REML = TRUE, control = lmerControl(),
          start = NULL, verbose = 0L, subset, weights, na.action,
          offset, contrasts = NULL, devFunOnly = FALSE, ...)
```

## Hypothesis testing?

With fixed effects, for a factor `f`, the comparison
```
anova( lm(y ~ f - 1), lm(y ~ 1) )
```
uses the model that
$$ y_i = \beta_{f_i} + \epsilon_i $$
to test against the null hypothesis that
$$ H_0 : \beta_1 = \beta_2 = \cdots = \beta_m = 0. $$

-----------

With *random* effects,
```
anova( lm(y ~ (1|f) - 1), lm(y ~ 1) )
```
uses the model that
$$\begin{aligned} 
    y_i &= \beta_{f_i} + \epsilon_i   \\
    \beta_a &\sim \Normal(0, \eta) 
\end{aligned}$$
to test against the null hypothesis that
$$ H_0 : \eta = 0. $$


# Back to the pumpkins

## Your turn

1. Add a *random effect* of `plot` to the model.
2. How big is the "plot" effect?
3. Assess significance by using `anova( )` to compare to a nested model.


Data: [data/pumpkins.tsv](data/pumpkins.tsv)
```{r read_pumpkins2}
pumpkins <- read.table("data/pumpkins.tsv", header=TRUE)
pumpkins$plot <- factor(pumpkins$plot)
pumpkins$fertilizer <- factor(pumpkins$fertilizer)
pumpkins$water <- factor(pumpkins$water)
```

## IN CLASS

**(1)** Add a random effect to the linear model.
```{r fit_pumpkin_model}
library(lme4)
fixed_fit <- lm( weight ~ water * fertilizer, data=pumpkins)
mixed_fit <- lmer( weight ~ water * fertilizer + (1|plot:water:fertilizer), data=pumpkins)
summary(mixed_fit)
```

------------

**(2)** How big is the "plot" effect?
The estimated standard deviation of the plot effects is
`r attr(summary(mixed_fit)$varcor[["plot:water:fertilizer"]], "stddev")`.
We can also look at the estimated random effects visually,
by plotting the estimated mean plot effects,
with a bar showing plus/minus 2 standard errors.
```{r plot_effect}
rfs <- ranef(mixed_fit)[["plot:water:fertilizer"]]
ses <- rfs[,1] + outer(sqrt(as.vector(attr(rfs, "postVar"))), c(-2, 2), "*")
plot(rfs[,1], 1:nrow(rfs), xlab='plot effect mean',
     xlim=range(ses), ylab='')
segments(x0=ses[,1], x1=ses[,2], y0=1:nrow(rfs))
abline(v=0, col='red')
```

----------------


Overall, our best guess is that plots differ by around 0.2 pounds,
but no more than 0.6 pounds from each other
(in terms of average pumpkin weight, with 95% confidence).

------------------

**(3)** Compare the models with/without random effects,
to see if the random effects seem to have added anything useful.

```{r compare_models}
anova(mixed_fit, fixed_fit)
```
Hm, a $p$-value of 1. That's suspicious. We know there *shouldn't* be an effect of plot,
but are things really working right?

---------------

See what happens if there *is* a plot effect

```{r new_sim}
sim_pumpkins <- function (plot_sd=0) {
    resid_sd <- 1.2
    plants_per_plot <- 5
    plots_per_treatment <- 4
    pumpkins <- expand.grid(
                  fertilizer=c("low", "medium", "high"),
                  water=c("no water", "water"),
                  plot=1:plots_per_treatment,
                  plant=1:plants_per_plot,
                  weight=NA)
    # true values
    mu <- 20
    alpha <- c('high'=0, 'medium'=-6, 'low'=-12)
    beta <- c('no water'=0, 'water'=0)
    gamma <- c('high.no water'=0,
               'high.water'=0,
               'medium.no water'=0,
               'medium.water'=2,
               'low.no water'=0,
               'low.water'=-2)
    plot_delta <- rnorm(plots_per_treatment
                        * nlevels(pumpkins$fertilizer)
                        * nlevels(pumpkins$water),
                        mean=0, sd=plot_sd)
    k <- 1
    for (p in 1:plots_per_treatment) {
        for (f in levels(pumpkins$fertilizer)) {
            for (w in levels(pumpkins$water)) {
                names(plot_delta)[k] <- paste(p, f, w, sep='.')
                k <- k + 1
            }
        }
    }
    pumpkins$mean_weight <- (mu
        + alpha[as.character(pumpkins$fertilizer)]
        + beta[as.character(pumpkins$water)]
        + gamma[paste(pumpkins$fertilizer, pumpkins$water, sep='.')]
        + plot_delta[paste(pumpkins$plot, pumpkins$fertilizer, pumpkins$water, sep='.')])
    pumpkins$weight <- rnorm(nrow(pumpkins),
                             mean=pumpkins$mean_weight,
                             sd=resid_sd)
    return(pumpkins)
}
```

---------------

```{r new_sim2}
new_pumpkins <- sim_pumpkins(plot_sd=3)
new_fixed_fit <- lm(
          weight ~ water * fertilizer, data=new_pumpkins)

new_mixed_fit <- lmer(
          weight ~ water * fertilizer + (1|plot:water:fertilizer),
          data=new_pumpkins)

anova(new_mixed_fit, new_fixed_fit)
```
Yes, everything seems to be working right!

---------------

In conclusion, there's no good evidence that we need a random effect for plots
(i.e., that plots are inducing unmodeled correlation in our data).
But it's not hurting our model to include them, either.

