---
title: "On ordination and dimension reduction methods"
author: "Peter Ralph"
date: "23 February 2020 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(lars)
library(tidyverse)
library(rstan)
library(matrixStats)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


# An ordination of dimension reduction techniques?

## The menagerie

There are *many* dimension reduction methods, e.g.:

- principal components analysis (PCA)
- non-negative matrix factorization (NMF)
- independent components analysis (ICA)
- canonical correpondence analysis (CCA)
- principal coordinates analysis (PCoA)
- multidimensional scaling (MDS)
- redundancy analysis (RDA)
- Sammon mapping
- kernel PCA
- t-SNE
- UMAP
- locally linear embedding (LLE)
- Laplacian eigenmaps
- autoencoders

## Using distances or similarities?

PCA uses the *covariance matrix*, which measures similarity.

t-SNE begins with the matrix of *distances*, measuring dissimilarity.

## Metric or non-Metric?

Are distances interpretable?

. . .

*metric:* In PCA, each axis is a fixed linear combination of variables.
So, distances always mean the same thing no matter where you are on the plot.

. . .

*non-metric:* In t-SNE,
distances within different clusters are not comparable.


## Why ordination?

From [ordination.okstate.edu](http://ordination.okstate.edu/overview.htm),
about ordination in ecology:


1. Graphical results often lead to intuitive interpretations of species-environment relationships.

2. A single multivariate analysis saves time, in contrast to a separate univariate analysis for each species.

3. Ideally and typically, dimensions of this 'low dimensional space' will represent important and interpretable environmental gradients.

4. If statistical tests are desired, problems of multiple comparisons are diminished when species composition is studied in its entirety.

5. Statistical power is enhanced when species are considered in aggregate, because of redundancy.

6. By focusing on 'important dimensions', we avoid interpreting (and misinterpreting) noise.



## Beware overinterpretation


1. Ordination methods
    are strongly influenced by *sampling*:
    ordination may ignore large-scale patterns in favor of describing variation within a highly oversampled area.

2. Ordination methods also describe patterns common to many variables:
   measuring the same thing many times may drive results.

3. Many methods are designed to find clusters, because our brain likes to categorize things.
   This doesn't mean those clusters are well-separated in reality.


## Some questions to ask

The goal is usually to produce a picture
in which similar things are nearby each other,
while also capturing global structure.

1. How is similarity measured in the original data?
2. How does the algorithm use that information?


# Text analysis

## Identifying authors

In [data/passages.txt](data/passages.txt)
we have a number of short passages from a few different books.

. . .

Can we identify the authors of each passage?

. . .

The true sources of the passages are in [data/passage_sources.tsv](data/passage_sources.tsv).

## Turn the data into a matrix

```{r wordmat, cache=TRUE}
passages <- readLines("data/passages.txt")
sources <- read.table("data/passage_sources.tsv", header=TRUE, stringsAsFactors=TRUE)
words <- sort(unique(strsplit(paste(passages, collapse=" "), " +")[[1]]))
tabwords <- function (x, w) { tabulate(match(strsplit(x, " ")[[1]], w), nbins=length(w)) }
wordmat <- sapply(passages, tabwords, words)
dimnames(wordmat) <- list(words, NULL)
stopifnot( min(rowSums(wordmat)) > 0 )
wordmat[1:20, 1:20]
```

## PCA?

```{r wordpca, fig.width=2.5*fig.dim, fig.height=1.5*fig.dim, cache=TRUE, dependson="wordmat"}
wordpcs <- prcomp(wordmat, scale.=TRUE)
layout(t(1:2))
plot(wordpcs$rotation[,1:2], col=sources$source, pch=20, xlab="PC1", ylab="PC2")
plot(wordpcs$rotation[,2:3], col=sources$source, pch=20, xlab="PC2", ylab="PC3")
legend("topright", pch=20, col=1:3, legend=levels(sources$source))
```

## PC1 is shortness

```{r wordlen, fig.width=2.0*fig.dim, fig.height=1.5*fig.dim}
plot(colSums(wordmat), wordpcs$rotation[,1], col=sources$source, xlab='length', ylab='PC1')
```

## PC2 is book

::: {.columns}
:::::::::::::: {.column width=50%}

```{r wordloadings, echo=FALSE}
xw <- wordpcs$x
head(xw[order(xw[,2]),1:3], 50)
```

:::
:::::::::::::: {.column width=50%}

```{r wordloadings2, echo=FALSE}
head(xw[order(xw[,2],decreasing=TRUE),1:3], 50)
```

:::
::::::::::::::


## PC3 ???

::: {.columns}
:::::::::::::: {.column width=50%}

```{r wordloadings3, echo=FALSE}
head(xw[order(xw[,3]),1:3], 50)
```

:::
:::::::::::::: {.column width=50%}

```{r wordloadings4, echo=FALSE}
head(xw[order(xw[,3],decreasing=TRUE),1:3], 50)
```

:::
::::::::::::::





# Visualizing expression space

## A conceptual model

Let's build a *conceptual* model for
descriptive analysis of "mixture" expression data.

. . .

**Data:** expression data from tissue samples
that consist of various *mixtures* of different cell types.

. . .

**Goal:** identify shared coexpression patterns
corresponding to *cell type*.

. . .

*Similar situations:* 
identify different developmental stages from whole-organism expression;
common community structures from metagenomic data.

----------------


1. Each cell type has a typical set of *mean* expression levels.

2. Each sample is composed of a mixture of cell types,
   defined by the proportions that come from each type.


------------------

::: {.columns}
:::::::::::::: {.column width=50%}


1. Mean expression by cell type.

2. Cell type proportions by sample.


:::
:::::::::::::: {.column width=50%}


1. $x_{kj}$ : Mean expression of gene $j$ in cell type $k$.

2. $w_{ik}$ : Proportion of sample $i$ of cell type $k$.

$Z_{ij}$ : expression level in sample $i$ of gene $j$.

   $$\begin{aligned}
        Z_{ij} \approx \sum_{k=1}^K w_{ik} x_{kj} .
   \end{aligned}$$


:::
:::::::::::::: 


