---
title: "Imputation of missing data"
author: "Peter Ralph"
date: "12 January 2021 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(brms)
library(rstan)
library(matrixStats)
library(tidyverse)
options(mc.cores = parallel::detectCores())
```



# Missing data

## 

Sometimes the data isn't all there.

What to do?

##

Options:

1. Restrict to non-missing observations.

2. Impute the missing data and fit a model.

3. Fit a model that also imputes the missing data.

 . . .

*Notes:*

- beware of *informative missingness*

- "impute" is a fancy word for "guess"


# A partially observed oscillator

##

The system itself does
$$\begin{aligned}
    x_{t+1} - x_t &= \alpha y_t + \Normal(0, \sigma_{xy}) \\
    y_{t+1} - y_t &= - \beta x_t + \Normal(0, \sigma_{xy}) 
\end{aligned}$$
but we only get to observe
$$\begin{aligned}
    X_t &= x_t + \Normal(0, \sigma_\epsilon) \\
    Y_t &= y_t + \Normal(0, \sigma_\epsilon) .
\end{aligned}$$



## Even more realism?

Now what if we actually *don't observe* most of the $Y$ values?

Here's what this looks like.
```{r sim_osc, cache=TRUE, fig.width=3*fig.dim, fig.height=1.5*fig.dim}
N <- 500
true_osc <- list(alpha=.1,
                 beta=.05,
                 sigma_xy=.05,
                 sigma_eps=.5)
xy <- matrix(nrow=N, ncol=2)
xy[1,] <- c(3,0)
for (k in 1:(N-1)) {
    xy[k+1,] <- (xy[k,] 
                + c(true_osc$alpha * xy[k,2],
                    (-1) * true_osc$beta * xy[k,1])
                + rnorm(2, 0, true_osc$sigma_xy))
}
XY <- xy + rnorm(N*2, 0, true_osc$sigma_eps)
obs_y <- sample.int(N, size=10)
XY[setdiff(1:N, obs_y), 2] <- NA
```

-----------------

```{r plot_osc, echo=FALSE}
matplot(xy, xlab='time', ylab='x and y', type='l')
matpoints(XY)
```

## A new Stan block

```{r osc_stan, cache=TRUE}
osc_block_missing <- "
data {
    int N;
    vector[N] X;
    int k; // number of observed Y
    int obs_y[k]; // which Y values are observed
    vector[k] Y;
}
parameters {
    real alpha;
    real beta;
    real<lower=0> sigma_xy;
    real<lower=0> sigma_eps;
    vector[N] x;
    vector[N] y;
}
model {
    x[1] ~ normal(0, 5);
    y[1] ~ normal(0, 5);
    x[2:N] ~ normal(x[1:(N-1)] + alpha * y[1:(N-1)], sigma_xy);
    y[2:N] ~ normal(y[1:(N-1)] - beta * x[1:(N-1)], sigma_xy);
    X ~ normal(x, sigma_eps);
    Y ~ normal(y[obs_y], sigma_eps);
    alpha ~ normal(0, 1);
    beta ~ normal(0, 1);
    sigma_xy ~ normal(0, 1);
    sigma_eps ~ normal(0, 1);
}
"
osc_model_missing <- stan_model(model_code=osc_block_missing)
```

-------------------

```{r run_osc, cache=TRUE, dependson=c("osc_stan", "sim_osc")}
osc_fit <- sampling(osc_model_missing,
                    data=list(N=N,
                              X=XY[,1],
                              k=length(obs_y),
                              obs_y=obs_y,
                              Y=XY[obs_y,2]),
                    iter=1000, chains=3,
                    control=list(max_treedepth=12))
```

## How'd we do?

```{r summarize_osc}
cbind(truth=c(true_osc$alpha, true_osc$beta, true_osc$sigma_xy, true_osc$sigma_eps),
      rstan::summary(osc_fit, pars=c("alpha", "beta", "sigma_xy", "sigma_eps"))$summary)
```

-------------------

Here is a density plot of 100 estimated trajectories (of `x` and `y`) from the Stan fit.

```{r show_osc_fit, echo=FALSE, fig.height=1.5*fig.dim, fig.width=1.5*fig.dim}
osc_results <- rstan::extract(osc_fit)
plot(xy, lwd=2, xlab='x', ylab='y', type='l')
for (k in 1:100) {
    lines(osc_results$x[k,], osc_results$y[k,],
          col=adjustcolor('black', 0.2))
}
```
