---
title: "Poisson linear models"
author: "Peter Ralph"
date: "Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(tidyverse)
library(brms)
library(bayesplot)
library(matrixStats)
options(mc.cores = parallel::detectCores())
```


# Count data

## A hypothetical situation:

1. We have **counts** of mutations,

2. from some individuals of different **ages**
   and past **exposures** to solar irradiation,

3. of two **genotypes**.

. . .

*Model:*

* Counts are **Poisson**,

* with mean that depends on age and exposure,

* but effect of exposure depends on genotype.



-------------------------

::: {.columns}
::::::::: {.column width="50%"}


1. Counts are **Poisson**,

2. with mean that depends on age and exposure,

3. but effect of exposure depends on genotype.

:::
:::::::::: {.column width="50%"}


$$\begin{aligned}
    Z_i &\sim \Poisson(\mu_i) \\
\end{aligned}$$

:::
:::::::::::



-------------------------

::: {.columns}
::::::::: {.column width="50%"}


1. Counts are **Poisson**,

2. with mean that depends on age and exposure,

3. but effect of exposure depends on genotype.

:::
:::::::::: {.column width="50%"}


$$\begin{aligned}
    Z_i &\sim \Poisson(\mu_i) \\
    \mu_i &= a + b \times \text{age}_i + c \times \text{exposure}_i 
\end{aligned}$$

:::
:::::::::::

-------------------------

::: {.columns}
::::::::: {.column width="50%"}


1. Counts are **Poisson**,

2. with mean that depends on age and exposure,

3. but effect of exposure depends on genotype.

:::
:::::::::: {.column width="50%"}


$$\begin{aligned}
    Z_i &\sim \Poisson(\mu_i) \\
    \mu_i &= a_{g_i} + b \times \text{age}_i + c_{g_i} \times \text{exposure}_i
\end{aligned}$$

:::
:::::::::::



-------------------------

::: {.columns}
::::::::: {.column width="50%"}


1. Counts are **Poisson**,

2. with mean that depends on age and exposure,

3. but effect of exposure depends on genotype.

:::
:::::::::: {.column width="50%"}


$$\begin{aligned}
    Z_i &\sim \Poisson(\mu_i) \\
    \mu_i &= \exp\left( a_{g_i} + b \times \text{age}_i \right. \\
        &\qquad \left. {} + c_{g_i} \times \text{exposure}_i \right)
\end{aligned}$$

:::
:::::::::::

# Poisson modeling, in practice

## The data


```{r sim_counts, include=FALSE, cache=TRUE}
true_params <- list(a=c(0, 2),
                    b=1/20,
                    c=c(1/10, -1/10),
                    sigma=0.7)
nsamples <- 500
data <- data.frame(genotype=factor(sample(c('A','B'), nsamples, 
                                          replace=TRUE)),
                   age = rgamma(nsamples, 10, 1),
                   exposure = rexp(nsamples, 0.2))
data$y <- with(data, true_params$a[genotype] +
                      true_params$b * age +
                      true_params$c[genotype] * exposure)
data$mu <- exp(rnorm(nrow(data), mean=data$y, 
                     sd=true_params$sigma))
data$counts <- rpois(nsamples, data$mu)
```

```{r save_data, include=FALSE}
write.table(data, file="data/poisson_counts_data.tsv", row.names=FALSE)
write.table(data[,c(1,2,3,6)], file="data/poisson_counts.tsv", row.names=FALSE)
```

```{r plot_counts, echo=FALSE, fig.width=3*fig.dim, fig.height=1.5*fig.dim}
layout(t(1:2))
plot(counts ~ age, col=genotype, data=data, pch=20)
for (g in levels(data$genotype)) {
    abline(coef(lm(counts ~ age, data=subset(data, genotype == g))), col=match(g, levels(data$genotype)))
}
plot(counts ~ exposure, col=genotype, data=data, pch=20)
for (g in unique(data$genotype)) {
    abline(coef(lm(counts ~ exposure, data=subset(data, genotype == g))), col=match(g, levels(data$genotype)))
}
legend("topright", pch=20, col=1:2,
       legend=paste("genotype", 1:2))
```

-----------


Here are [the data](data/poisson_counts.tsv):
```{r show_data}
head( countdata <- read.table("data/poisson_counts.tsv", header=TRUE, stringsAsFactors=TRUE) )
```


## A GLM

```{r pois_fmla, cache=TRUE}
bf <- brmsformula(counts ~ 0 + genotype + age + genotype:exposure)
get_prior(bf, data=countdata)
```

-----------

```{r run_simple_pois, cache=TRUE, dependson=c("sim_counts", "pois_fmla")}
fit1 <- brm(
   counts ~ 0 + genotype + age + genotype:exposure,
   family=poisson(link='log'),
   prior=set_prior("normal(0, 1)", class="b"),
   data=countdata
)
```

## 

```{r printit}
print(fit1)
```

## 

```{r not_warmup}
mcmc_trace(fit1, regex_pars=c("b_.*"))
```

## 

Posterior distributions of the parameters:
```{r true_fit_1, echo=FALSE}
mcmc_intervals(fit1, regex_pars=c("b_.*"))
```


# Goodness of fit

## Posterior predictive simulations

*Idea:* Let's simulate up data *under this model*
and see if it looks like the real data.

. . .

A bunch of times, we'll

1. pick a set of parameters from the posterior
2. compute the vector of means (`mu`)
3. simulate a vector of Poisson counts with mean `mu`

. . .

... with the `pp_check` function!

## The data are *overdispersed* relative to our simulations

```{r plot_post_sims1, echo=-1, fig.height=1.5*fig.dim}
fit1$data$order <- rank(countdata$counts, ties='first')
pp_check(fit1, type='intervals', x="order")
```

## One solution:

Use the `negbinomial ` family

. . .

... which introduces more randomness:

$$\begin{aligned}
    Z_i &\sim \Poisson(M_i) \\
    M_i &\sim \Gam(\text{mean}=\mu_i, \text{shape}=r) \\
    \mu_i &= \exp(y_i) \\
    y_i &= a_{g_i} + b \times \text{age}_i \\
        &\qquad {} + c_{g_i} \times \text{exposure}_i
\end{aligned}$$


-----------

```{r run2, cache=TRUE, dependson=c("sim_counts", "pois_fmla")}
fit2 <- brm(
   counts ~ 0 + genotype + age + genotype:exposure,
   family=negbinomial(link='log'),
   prior=set_prior("normal(0, 1)", class="b"),
   data=countdata
)
```
## 

```{r printit2}
print(fit2)
```

## 

```{r not_warmup2}
mcmc_trace(fit2, regex_pars=c("b_.*"))
```

## 

Posterior distributions of the parameters:
```{r true_fit_2, echo=FALSE}
mcmc_intervals(fit2, regex_pars=c("b_.*"))
```

---------------


```{r plot_post_sims2, echo=-1, fig.height=1.5*fig.dim}
fit2$data$order <- rank(countdata$counts, ties='first')
pp_check(fit2, type='intervals', x="order")
```


