---
title: "Parametric Survival Analysis"
author: "Peter Ralph"
date: "5 January 2021 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center',
                      warning=FALSE)
set.seed(42)
library(brms)
library(survival)
library(rstan)
library(matrixStats)
options(mc.cores = parallel::detectCores())
```



# Fully parametric models

## Modeling

The idea: choose a (parametric) distribution for the time-until-event.

. . .

Some choices:

- Exponential (constant rate)
- Weibull
- Gompertz
- logNormal

##

::: {.columns}
:::::: {.column width=50%}

Model:

- lifetimes are Weibull distributed
- mean lifetime depends on covariates

:::
:::::: {.column width=50%}

:::
::::::

##

::: {.columns}
:::::: {.column width=50%}

Model:

- lifetimes are Weibull distributed
- mean lifetime depends on covariates

:::
:::::: {.column width=50%}

$$\begin{aligned}
    T_i &\sim \Weibull(\text{shape}=k, \text{mean}=\mu_i) \\
    \mu_i &= \exp(y_i) \\
    y_i &= \beta_0 + \sum_j \beta_j x_{ij} 
\end{aligned}$$

:::
::::::

## 

Three ways to do parametric survival analysis:

- `survival::survreg` : maximum likelihood
- `brms::brm` : Baysian, with Stan
- MC Stan

## Maximum likelihood: survreg

```{r survreg}
weifit <- survreg(Surv(time, status) ~ sex + ph.ecog + ph.karno,
                  data = lung, dist='weibull')
```

## Bayesian: brms

```{r brms, cache=TRUE}
lung$censored <- ifelse(lung$status == 1, 'right', 'none')

brmfit <- brm(time | cens(censored) ~ sex + ph.ecog + ph.karno,
              data=lung, family='weibull')
```

##

```{r dummy, include=FALSE}
summary(weifit)
summary(brmfit)
```

::: {.columns}
:::::: {.column width=50%}

<pre style="width:120%; margin-left:-20%;">
```{r ssum, echo=FALSE, results='asis', message=FALSE}
summary(weifit)
```
</pre>

:::
:::::: {.column width=50%}

<pre style="width:120%;">
```{r bsum, echo=FALSE, results='asis', message=FALSE}
summary(brmfit)
```
</pre>

:::
::::::


## Random effects, with brms

```{r brms2, cache=TRUE, echo=2:3}
lung$inst[is.na(lung$inst)] <- 100
brmfit2 <- brm(time | cens(censored) ~ sex + ph.ecog + ph.karno + (1|inst),
               data=lung, family='weibull')
```

##

```{r sb2, message=FALSE}
summary(brmfit2)
```

##

```{r insts}
ranef(brmfit2)
```


# Application

## 

Fit (and select) a proportional hazards model to the `veteran` dataset.
What are the most important predictors of survival?

```{r vetdata, eval=1}
data(veteran)
help(veteran)
```
```
Veterans' Administration Lung Cancer study

Description:

     Randomised trial of two treatment regimens for lung cancer.  This
     is a standard survival analysis data set.

Usage:

     veteran
     
Format:

       trt:       1=standard 2=test                            
       celltype:  1=squamous,  2=smallcell,  3=adeno,  4=large 
       time:      survival time                                
       status:    censoring status                             
       karno:     Karnofsky performance score (100=good)       
       diagtime:  months from diagnosis to randomisation       
       age:       in years                                     
       prior:     prior therapy 0=no, 10=yes      
```


# Review

##

Nonparametric:

- simple
- does not depend on model fit

. . .

Parametric:

- makes more sense: real hazard rates are not piecewise constant
- better power, *if* the model fits
- can simulate



## The Kaplan-Meier curve, by hand:

```{r byhand}
f1 <- survfit(Surv(time, status) ~ 1, data = lung)

y <- data.frame(time=sort(unique(lung$time[lung$status == 2])))
y$k <- sapply(y$time, function (t) sum(lung$status == 2 & lung$time == t))
y$n <- sapply(y$time, function (t) sum(lung$time > t | (lung$time == t & lung$status == 2)))
y$p <- cumprod(1 - y$k/y$n)

plot(f1)
points(p ~ time, data=y, col='red', pch=20, cex=0.5)
```
