---
title: "Untitled"
author: "Peter Ralph"
date: "October 5, 2021"
output: html_document
---

Read in data:
```{r data}
airbnb <- read.csv("../Datasets/Airbnb_data/Airbnb_listings_Portland.csv")
prices <- na.omit(as.numeric(gsub("$", "", airbnb$price, fixed=TRUE)))
```

Find the difference between sample mean and true mean as a function of $n$:
```{r diff_mean}
true_mean <- mean(prices)
mean_diff <- function (n) {
    # This function takes a sample size, n, and estimates
    # the mean absolute error that you get when estimating
    # the mean price using a sample of size n.
    differences <- replicate(200, {
        x <- sample(prices, size=n)
        abs(mean(x) - true_mean)  # absolute error
    })
    return(mean(differences)) # mean absolute error
}
```
Let's make a plot of how the mean absolute error depends on $n$:
```{r plotit}
nvals <- 10*(10:100)
maes <- rep(NA, length(nvals))
for (j in seq_along(nvals)) {
    maes[j] = mean_diff(nvals[j])
}
plot(nvals, maes, type='l',
     xlab='sample size (n)',
     ylab='mean absolute error ($)')
```
We see that, for instance, if we take a sample of size 200,
then we'll probably get an estimate of the mean within around $7
of the true mean,
and at a sample of size 1000, the error will be between 2 and 3 dollars.

## When is the error within $10?

Next we'd like to estimate the *probability* that a sample
of size $n$ will return an estimate within $10 of the true mean.
```{r prob_mean}
prop_lt_10 <- function (n) {
    # This function takes a sample size, n, and estimates
    # the probability that the sample mean is within $10
    # of the true mean.
    differences <- replicate(200, {
        x <- sample(prices, size=n)
        abs(mean(x) - true_mean)  # absolute error
    })
    return(mean(differences < 10)) # proportion less than 10
}
```
Here's the results:
```{r plot_probs}
probs <- sapply(nvals, prop_lt_10)
plot(nvals, probs, type='l',
     xlab='sample size (n)',
     ylab='prob error is less than $10')
```
To be confident that the error is within $10,
it looks like you'd need a sample of at least size 400 or 500.