---
title: "The Poisson distribution"
author: "Peter Ralph"
date: "Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(tidyverse)
library(rstan)
library(matrixStats)
options(mc.cores = parallel::detectCores())
```



# Stochastic minute

## The Poisson distribution

If $N \sim \Poisson(\mu)$ then $N \ge 0$ and
$$\begin{aligned}
    \P\{N = k\} = \frac{\mu^k}{k!} e^{-\mu}
\end{aligned}$$

> - $N$ is a nonnegative integer (i.e., a *count*)
> 
> - $\E[N] = \var[N] = \mu$
> 
> - If a machine makes widgets very fast,
>   producing on average one broken widget per minute (and many good ones),
>   each breaking independent of the others,
>   then the number of broken widgets in $\mu$ minutes is $\Poisson(\mu)$.
> 
> - If busses arrive randomly every $\Exp(1)$ minutes,
>   then the number of busses to arrive in $\mu$ minutes
>   is $\Poisson(\mu)$.

##

Important point:

- the standard deviation is the square root of the mean

```{r plot_poisson, echo=FALSE, fig.height=2*fig.dim, fig.width=2.5*fig.dim}
muvals <- c(1, 10, 100, 1000)
layout(matrix(1:4, nrow=2))
for (mu in muvals) {
    x <- rpois(1e5, mu)
    hist(x, xlim=range(0, x), xlab='value', main=sprintf("mean = %d", mu), freq=FALSE)
}
```
