---
title: "Formulas in R"
author: "Peter Ralph"
date: "Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(matrixStats)
library(tidyverse)
```

# Set-up

## Example data

Continuing with the "pumpkin data":

```{r get_data}
(pumpkins <- read.table("data/pumpkins.tsv"))
```

# Formulas

## Anatomy of a formula

```
  weight ~ fertilizer + water
```

is read something like

> mean weight is determined by additive effects of fertlizer and of water

. . .

$$\begin{equation}\begin{split}
    y &\sim x \qquad \text{means} \\
    y &= a + b x + \text{(mean-zero noise)} .
\end{split}\end{equation}$$

## Intercepts

The intercept is included implicitly, so these are equivalent:
```
  weight ~ fertilizer + water
  weight ~ 1 + fertilizer + water
```

. . .

... so if you *don't* want an intercept, do
```
  weight ~ 0 + fertilizer + water
```

## Interactions

To assign an effect to each element of a *crossed* design, use `:`, e.g.
```
  weight ~ fertilizer + water + fertilizer:water
```

. . .

which is the same as
```
  weight ~ fertilizer * water
```
since lower-order effects are included implicitly.


## A translation table

- `~` : depends on
- `+` : and also, independently
- `:` : in combination with
- `*` : and also
- `I(x+y)` : actually `x` plus `y`
- `I(x^2)` : actually `x` squared

. . .

Trickier things:

- `1` : an intercept
- `0` : but not an intercept
- `-` : but not
- `.` : all columns not otherwise in the formula
- `x/y` : `x`, and `y` nested within `x` (same as `x + x:y`)

## The secret to formulas



. . .

If you want to know what a formula is *really* doing,
look at its `model.matrix( )`,
whose columns correspond to the coefficients of the resulting model.

```{r model_matrix}
model.matrix(~ fertilizer, data=pumpkins)
```

-----------------

```{r plot_model_matrix}
matplot(
    model.matrix(~ fertilizer, data=pumpkins),
    type='l')
```

------------

```{r model_matrix2}
model.matrix(~ fertilizer + water, data=pumpkins)
```


------------

```{r plot_model_matrix2}
matplot(
    model.matrix(~ fertilizer + water, data=pumpkins),
    type='l')
```

------------

```{r model_matrix3, echo=2}
options(width=200)
model.matrix(~ 0 + fertilizer * water, data=pumpkins)
```

------------

```{r plot_model_matrix3}
matplot(
    model.matrix(~ 0 + fertilizer * water, data=pumpkins),
    type='l')
```

## Note:

For fine control of factors in linear models, either

- `relevel()` them, or
- manually set their `contrasts()`.

## Exercise:

Make formulas that give you estimates of

1. A global mean ($\mu$), two fertilizer effects ($\alpha_\text{medium}$ and $\alpha_\text{high}$),
    and one water effect ($\beta_\text{water}$).

2. Three fertilizer effects ($\alpha_\text{low}$, $\alpha_\text{medium}$ and $\alpha_\text{high}$),
    and one water effect ($\beta_\text{water}$).

3. Two fertilizer effects ($\alpha_\text{medium}$ and $\alpha_\text{high}$),
    and two water effect ($\beta_\text{no water}$ and $\beta_\text{water}$).

4. A single mean for each of the six conditions 
   ($\gamma_\text{high, water}$, $\gamma_\text{medium, water}$, $\gamma_\text{low, water}$, 
   $\gamma_\text{high, no water}$, $\gamma_\text{medium, no water}$, $\gamma_\text{low, no water}$).

## Example:

1. A global mean ($\mu$), two fertilizer effects ($\alpha_\text{medium}$ and $\alpha_\text{high}$),
    and one water effect ($\gamma_\text{water}$).

*Example:*
```{r lm1}
summary(lm(weight ~ fertilizer + water, data=pumpkins))
```
