---
title: "Categorical data with brms"
author: "Peter Ralph"
date: "Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(tidyverse)
library(rstan)
library(brms)
library(matrixStats)
options(mc.cores = parallel::detectCores())
options(digits=2)
```

```{r get_post_fn, echo=FALSE}
get_post_props <- function (fit, numer_names, denom_names) {
    # helper function to get posterior distribution of conditional proportions
    # from poisson(link='log') fits
    ppe <- posterior_epred(fit)
    prop_names <- do.call(expand.grid,
                          lapply(c(numer_names, denom_names), function (vn) levels(factor(fit$data[[vn]]))))
    names(prop_names) <- c(numer_names, denom_names)
    numer <- denom <- matrix(0, nrow=nrow(ppe), ncol=nrow(prop_names))
    for (k in 1:nrow(fit$data)) {
        dyes <- rep(TRUE, nrow(prop_names))
        for (vn in denom_names) {
            dyes <- dyes & (prop_names[[vn]] == fit$data[[vn]][k])
        }
        nyes <- dyes
        for (vn in numer_names) {
            nyes <- nyes & (prop_names[[vn]] == fit$data[[vn]][k])
        }
        if (sum(nyes) > 0) {
            numer[,nyes] <- numer[,nyes] + ppe[,k]
        }
        if (sum(dyes) > 0) {
            denom[,dyes] <- denom[,dyes] + ppe[,k]
        }
    }
    post_props <- reshape2::melt(numer / denom)
    for (vn in c(numer_names, denom_names)) {
        post_props[[vn]] <- prop_names[[vn]][post_props$Var2]
    }
    return(post_props)
}
```



# Categorical data

## Hair and Eye color

```{r hair_eye_data}
data(HairEyeColor)
```
```
HairEyeColor             package:datasets              R Documentation

Hair and Eye Color of Statistics Students

Description:

     Distribution of hair and eye color and sex in 592 statistics
     students.

Usage:

     HairEyeColor
     
Format:

     A 3-dimensional array resulting from cross-tabulating 592
     observations on 3 variables.  The variables and their levels are
     as follows:

       No  Name  Levels                    
        1  Hair  Black, Brown, Red, Blond  
        2  Eye   Brown, Blue, Hazel, Green 
        3  Sex   Male, Female              
      
Details:

     The Hair x Eye table comes from a survey of students at the
     University of Delaware reported by Snee (1974).  The split by
     ‘Sex’ was added by Friendly (1992a) for didactic purposes.

     This data set is useful for illustrating various techniques for
     the analysis of contingency tables, such as the standard
     chi-squared test or, more generally, log-linear modelling, and
     graphical methods such as mosaic plots, sieve diagrams or
     association plots.

Source:

     <URL:
     http://euclid.psych.yorku.ca/ftp/sas/vcd/catdata/haireye.sas>

     Snee (1974) gives the two-way table aggregated over ‘Sex’.  The
     ‘Sex’ split of the ‘Brown hair, Brown eye’ cell was changed to
     agree with that used by Friendly (2000).

References:

     Snee, R. D. (1974).  Graphical display of two-way contingency
     tables.  _The American Statistician_, *28*, 9-12.  doi:
     10.2307/2683520 (URL: http://doi.org/10.2307/2683520).

```

----------------

```{r show_he, echo=FALSE}
HairEyeColor
```

---------------

```{r melt_he}
haireye <- as.data.frame(HairEyeColor)
names(haireye) <- tolower(names(haireye))
names(haireye)[names(haireye) == "freq"] <- "number"
haireye
```

## 

*Questions:* 

1. Are hair and eye color independent in this sample?
2. Do hair and eye color proportions differ by gender?


# Poisson models for count data

##

We have for data
$$\begin{aligned}
    n_{ij} = \text{( number of observations in categories $i$ and $j$ )} .
\end{aligned}$$
and
$$\begin{aligned}
    n_{i \cdot} &= \text{( total number of observations in category $i$ )} , \\
    n_{\cdot j} &= \text{( total number of observations in category $j$ )} , \\
    n_{\cdot \cdot} &= \text{( total number of observations )} .
\end{aligned}$$

. . .

*Question:* Is $n_{i \cdot}$ fixed? What about $n_{\cdot j}$? Or $n_{\cdot \cdot}$?

. . .

*Answer:* Happily, it will turn out not to matter (for the model).

## Poisson additivity

If we have Poisson-many things of two categories:
$$\begin{aligned}
    A &\sim \Poisson(a) \\
    B &\sim \Poisson(b)
\end{aligned}$$
then the total number of things is also Poisson:
$$
    A + B \sim \Poisson(a + b)
$$
and each thing chooses its type independently:
$$
    A \given A + B \sim \Binom\left(A+B, \frac{a}{a+b}\right) .
$$

## Poisson models for contingency tables

So, if we fit a model where 

$$\begin{aligned}
    n_{ij}
    &\sim \Poisson(\lambda_{ij})
\end{aligned}$$

then 
$$\begin{aligned}
    \frac{\lambda_{ij}}{\sum_{k\ell} \lambda_{k\ell}}
    &=
    (\text{proportion in category $ij$}) \\
    \frac{\lambda_{ij}}{\sum_{k} \lambda_{kj}}
    &=
    (\text{proportion of row $j$ in column $i$}) \\
    \frac{\lambda_{ij}}{\lambda_{ij} + \lambda_{k\ell}}
    &=
    (\text{proportion of those in $k\ell$ or $ij$ that are $ij$})
\end{aligned}$$

-------------------

For example, the *proportion of black-haired people who have blue eyes* is
$$\begin{aligned}
   & \P\left(\text{ eye blue } \given \text{ hair black } \right) \\
   &\qquad =
   \frac{
    \lambda_{\text{blue},\text{black}}
   }{
    \lambda_{\cdot,\text{black}}
   } \\
   &\qquad =
   \frac{
    \lambda_{\text{blue},\text{black}}
   }{
    \lambda_{\text{brown},\text{black}}
    + \lambda_{\text{blue},\text{black}}
    + \lambda_{\text{hazel},\text{black}}
    + \lambda_{\text{green},\text{black}}
   }
\end{aligned}$$

. . . 

*Question:* What's
$$\begin{aligned}
   \P\left(\text{ male } \given \text{ hair black } \right) = ?
\end{aligned}$$



## Independence and multiplicativity

If hair and eye color are *independent*,
then probabilities of combinations are *multiplicative*:

$$\begin{aligned}
    &\P\{\text{black hair and blue eyes}\} \\
    &\qquad =
        \P\{\text{black hair}\} \times \P\{\text{blue eyes}\given\text{black hair}\} \\
\end{aligned}$$

. . .

which if independent is
$$\begin{aligned}
    &\hphantom{\P\{\text{black hair and blue eyes}\}} \\
    &\qquad =
        \P\{\text{black hair}\} \times \P\{\text{blue eyes}\}
\end{aligned}$$

. . .

In shorthand,
$$\begin{aligned}
   p_{ij} = p_i \times p_j . 
\end{aligned}$$

## Additivity and independence

If we use a *log link function*, additivity corresponds to independence:
$$\begin{aligned}
    p_{ij}
    &=
        \frac{
        \lambda_{ij}
        }{
        \sum_{ab} \lambda_{ab}
        } \\
    &=
        \frac{
        e^{\beta_i + \gamma_j}
        }{
        \sum_{ab} e^{\beta_a + \gamma_b}
        } \\
    &=
    \left(
        \frac{
        e^{\beta_i}
        }{
        \sum_{a} e^{\beta_a}
        }
    \right)
    \times
    \left(
        \frac{
        e^{\gamma_j}
        }{
        \sum_{b} e^{\gamma_b}
        }
    \right) \\
    &= p_i \times p_j 
\end{aligned}$$


# hair and eye colors, with brms


## brms

```{r brm_fmla, cache=TRUE}
he_formula <- brmsformula(number ~ sex + hair * eye + (hair + eye) * sex)
```

## Flat priors

```{r brm_prior, cache=TRUE, dependson='brm_fmla'}
get_prior(he_formula, data=haireye, family='poisson')
he_priors <- c(prior('normal(0,3)', class='b'))
```

##

```{r brm_fit, cache=TRUE, dependson='brm_prior'}
he_fit <- brm(he_formula,
              data=haireye,
              family=poisson(link='log'),
              prior=he_priors)
```

##

```{r brm_summ}
summary(he_fit)
```

##

```{r brm_results, fig.width=2.5*fig.dim, fig.height=2*fig.dim}
conditional_effects(he_fit, effects="hair:eye")
```

##

But, that was *numbers*:
here's the credible intervals for the proportions of each eye color for each hair color:
```{r brm_pred, echo=FALSE, fig.width=2.5*fig.dim, fig.height=2*fig.dim}
get_post_props(he_fit, "eye", "hair") %>% group_by(hair, eye) %>% 
    summarise(proportion=mean(value), q025=quantile(value, probs=0.025), q975=quantile(value, probs=0.975)) %>%
    ggplot(aes(x=hair, y=proportion, ymin=q025, ymax=q975, col=eye)) + geom_point(position=position_dodge(width=0.4), size=4) + geom_errorbar(aes(ymin=q025, ymax=q975), position=position_dodge(width=0.4), width=0.3)
```

##

Recall the "reference" levels are
hair = `r levels(haireye$hair)[1]`
and 
eye = `r levels(haireye$eye)[1]`
```{r coef_plot, fig.width=2.5*fig.dim, fig.height=1.5*fig.dim, echo=FALSE}
mcmc_plot(he_fit, variable="b_hair.*:eye.*", regex=TRUE)
```


# Back to the questions

## 

*Questions:* 

1. Are hair and eye color independent in this sample?
2. Do hair and eye color proportions differ by gender?

## *Are hair and eye color independent in this sample?*

(What do we mean: given gender? averaged across gender?)

Recall the "reference" levels are
hair = `r levels(haireye$hair)[1]`
and 
eye = `r levels(haireye$eye)[1]`
```{r bysexq, echo=FALSE}
mcmc_plot(he_fit, variable="b_hair.*:eye.*", regex=TRUE)
```

------------

Eye color proportions by hair type:
```{r brm_pred2, echo=FALSE, fig.width=2.5*fig.dim, fig.height=2*fig.dim}
get_post_props(he_fit, "eye", "hair") %>% group_by(hair, eye) %>% 
    summarise(proportion=mean(value), q025=quantile(value, probs=0.025), q975=quantile(value, probs=0.975)) %>%
    ggplot(aes(x=hair, y=proportion, ymin=q025, ymax=q975, col=eye)) + geom_point(position=position_dodge(width=0.4), size=4) + geom_errorbar(aes(ymin=q025, ymax=q975), position=position_dodge(width=0.4), width=0.3)
```

-------------

Hair color proportions by eye type:
```{r brm_pred3, echo=FALSE, fig.width=2.5*fig.dim, fig.height=2*fig.dim}
get_post_props(he_fit, "hair", "eye") %>% group_by(hair, eye) %>% 
    summarise(proportion=mean(value), q025=quantile(value, probs=0.025), q975=quantile(value, probs=0.975)) %>%
    ggplot(aes(x=eye, y=proportion, ymin=q025, ymax=q975, col=hair)) + geom_point(position=position_dodge(width=0.4), size=4) + geom_errorbar(aes(ymin=q025, ymax=q975), position=position_dodge(width=0.4), width=0.3)
```



## *Do hair and eye color proportions differ by gender?*

```{r bysex1, echo=FALSE}
# conditional_effects(he_fit, effects="eye:sex")
get_post_props(he_fit, "eye", "sex") %>% group_by(eye, sex) %>% 
    summarise(proportion=mean(value), q025=quantile(value, probs=0.025), q975=quantile(value, probs=0.975)) %>%
    ggplot(aes(x=sex, y=proportion, ymin=q025, ymax=q975, col=eye)) + geom_point(position=position_dodge(width=0.4), size=4) + geom_errorbar(aes(ymin=q025, ymax=q975), position=position_dodge(width=0.4), width=0.3)
```

------------

```{r bysex2, echo=FALSE}
# conditional_effects(he_fit, effects="hair:sex")
get_post_props(he_fit, "hair", "sex") %>% group_by(hair, sex) %>% 
    summarise(proportion=mean(value), q025=quantile(value, probs=0.025), q975=quantile(value, probs=0.975)) %>%
    ggplot(aes(x=sex, y=proportion, ymin=q025, ymax=q975, col=hair)) + geom_point(position=position_dodge(width=0.4), size=4) + geom_errorbar(aes(ymin=q025, ymax=q975), position=position_dodge(width=0.4), width=0.3)
```


# Testing our method

##

What we've done so far:

1. Fit the model,
2. observed CIs for interactions do not overlap zero
3. concluded good statistical support for nonindependence.

. . .

*Question:*
What would we have seen if hair and eye color *were* independent?


## Testing

Let's simulate some test data, under independence.
```{r sim_data, cache=TRUE}
sim_data <- function () {
    n <- sum(haireye$number)
    p_hair <- tapply(haireye$number, haireye$hair, sum) / n
    p_eye <- tapply(haireye$number, haireye$eye, sum) / n
    p_sex <- tapply(haireye$number, haireye$sex, sum) / n
    p <- (p_hair[haireye$hair] * 
          p_eye[haireye$eye] * 
          p_sex[haireye$sex])
    stopifnot(sum(p) == 1)
    number <- rmultinom(1, size=n, prob=p)
    return(number)
}
```

-------------

```{r sim_fit, cache=TRUE, dependson='sim_data', warning=FALSE}
sim_fits <- lapply(1:5, function (k) {
                       # replaces the data only locally
                       haireye$number <- sim_data()
                       update(he_fit, newdata=haireye) })
```

-------------


```{r plot_sims, fig.width=4*fig.dim, fig.height=3*fig.dim, echo=FALSE}
cowplot::plot_grid(plotlist=lapply(lapply(
                                 lapply(c(list(he_fit), sim_fits), conditional_effects, effects="hair:eye:"),
                             plot, plot=FALSE), "[[", 1),
                   nrow=2, ncol=3
)
```

-------------

```{r plot_sims2, fig.width=4*fig.dim, fig.height=3*fig.dim, echo=FALSE}
cowplot::plot_grid(
           plotlist=lapply(c(list(he_fit), sim_fits), mcmc_plot, variable="b_hair.*:eye.*", regex=TRUE),
           nrow=2, ncol=3)
```
