---
title: "Parametric Survival Analysis"
author: "Peter Ralph"
date: "Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center',
                      warning=FALSE)
set.seed(42)
library(brms)
library(survival)
library(rstan)
library(matrixStats)
options(mc.cores = parallel::detectCores())
```



# Fully parametric models

## Modeling

The idea: choose a (parametric) distribution for the time-until-event.

. . .

Some choices:

- Exponential (constant rate)
- Weibull
- Gompertz
- logNormal

##

::: {.columns}
:::::: {.column width=50%}

Model:

- lifetimes are Weibull distributed
- mean lifetime depends on covariates

:::
:::::: {.column width=50%}

:::
::::::

##

::: {.columns}
:::::: {.column width=50%}

Model:

- lifetimes are Weibull distributed
- mean lifetime depends on covariates

:::
:::::: {.column width=50%}

$$\begin{aligned}
    T_i &\sim \Weibull(\text{shape}=k, \\
        &\qquad \text{mean}=\mu_i) \\
    \mu_i &= \exp(y_i) \\
    y_i &= \beta_0 + \sum_j \beta_j x_{ij} 
\end{aligned}$$

:::
::::::

## 

Three ways to do parametric survival analysis:

- `survival::survreg` : maximum likelihood
- `brms::brm` : Baysian, with Stan
- MC Stan

## Maximum likelihood: survreg

```{r survreg}
weifit <- survreg(Surv(time, status) ~ sex + ph.ecog + ph.karno,
                  data = lung, dist='weibull')
```

## Bayesian: brms

```{r brms, cache=TRUE}
lung$censored <- ifelse(lung$status == 1, 'right', 'none')

brmfit <- brm(time | cens(censored) ~ sex + ph.ecog + ph.karno,
              data=lung, family='weibull')
```

##

```{r dummy, include=FALSE}
summary(weifit)
summary(brmfit)
```

::: {.columns}
:::::: {.column width=50%}

<pre style="width:120%; margin-left:-20%;">
```{r ssum, echo=FALSE, results='asis', message=FALSE}
summary(weifit)
```
</pre>

:::
:::::: {.column width=50%}

<pre style="width:120%;">
```{r bsum, echo=FALSE, results='asis', message=FALSE}
summary(brmfit)
```
</pre>

:::
::::::


## Random effects, with brms

Recall there are 19 different institutions:
```{r do_insts}
table( lung$inst <- factor(lung$inst) )
```

```{r brms2, cache=TRUE}
brmfit2 <- brm(time | cens(censored) ~ sex + ph.ecog + ph.karno + (1|inst),
               data=subset(lung, !is.na(inst)), family='weibull')
```

##

```{r sb2, message=FALSE}
summary(brmfit2)
```

-------------

The conditional effect of sex on mean survival time (with mean values of other variables):
```{r insts_sex}
conditional_effects(brmfit2, effects='sex')
```

-------------

The conditional effect of ECOG on mean survival time:
```{r insts_ecog}
conditional_effects(brmfit2, effects='ph.ecog')
```


-------------

The conditional effect of Karnofsky score on mean survival time:
```{r insts_karno}
conditional_effects(brmfit2, effects='ph.karno')
```

-------------

A posterior predictive check: data are dark points, posterior medians are large points,
and lines give 50% and 90% posterior ranges:
```{r ppcheck, message=FALSE}
pp_check(brmfit2, type='intervals')
```

---------------

Consulting [vignette("brms_families")](https://cran.r-project.org/web/packages/brms/vignettes/brms_families.html),
the survival curve is $\exp(-(t / \lambda)^k)$ where $\lambda = \mu / \Gamma(1 + 1/k)$. Here are 100 samples from the posterior:
```{r surv_curvs, echo=FALSE, fig.width=3*fig.dim, fig.height=2*fig.dim}
post_means <- posterior_epred(brmfit2, newdata=data.frame(age=60, sex=2, ph.ecog=0:3, ph.karno=50), re_formula=NA, dpar="mu")[1:100,]
post_shape <- posterior_epred(brmfit2, newdata=data.frame(age=60, sex=2, ph.ecog=0:3, ph.karno=50), re_formula=NA, dpar="shape")[1:100,]
tvals <- seq(0, 2000, length.out=101)
Svals <- array(NA, dim=c(length(tvals), nrow(post_means), ncol(post_means)))
for (k in 1:nrow(post_means)) {
    lambda <- post_means[k,] / gamma(1 + 1/post_shape[k,])
    for (j in 1:ncol(post_means)) {
        Svals[,k,j] <- exp(- (tvals / lambda[j])^post_shape[k,j])
    }
}
layout(matrix(1:4, nrow=2, byrow=TRUE))
for (j in 1:dim(Svals)[3]) {
    plot(tvals, rowMeans(Svals[,,j]), lty=1, ylim=c(0,1), type='l', xlab='time (t)', ylab='S(t)', col=j, main=paste("ECOG =", j-1))
    matlines(tvals, Svals[,,j], type='l', col=adjustcolor(j, 0.25), lty=1)
    abline(v=1000, lty=3)
}
```

----------

Here's the code for that last one:
```{r surv_curvs2, eval=FALSE}
post_means <- posterior_epred(brmfit2, newdata=data.frame(age=60, sex=2, ph.ecog=0:3, ph.karno=50), re_formula=NA, dpar="mu")[1:100,]
post_shape <- posterior_epred(brmfit2, newdata=data.frame(age=60, sex=2, ph.ecog=0:3, ph.karno=50), re_formula=NA, dpar="shape")[1:100,]
tvals <- seq(0, 2000, length.out=101)
Svals <- array(NA, dim=c(length(tvals), nrow(post_means), ncol(post_means)))
for (k in 1:nrow(post_means)) {
    lambda <- post_means[k,] / gamma(1 + 1/post_shape[k,])
    for (j in 1:ncol(post_means)) {
        Svals[,k,j] <- exp(- (tvals / lambda[j])^post_shape[k,j])
    }
}
layout(matrix(1:4, nrow=2, byrow=TRUE))
for (j in 1:dim(Svals)[3]) {
    plot(tvals, rowMeans(Svals[,,j]), lty=1, ylim=c(0,1), type='l', xlab='time (t)', ylab='S(t)', col=j, main=paste("ECOG =", j-1))
    matlines(tvals, Svals[,,j], type='l', col=adjustcolor(j, 0.25), lty=1)
    abline(v=1000, lty=3)
}
```


# Application

## 

Fit (and select) a proportional hazards model to the `veteran` dataset.
What are the most important predictors of survival?

```{r vetdata, eval=1}
data(cancer, package='survival')
help(veteran)
```
```
Veterans' Administration Lung Cancer study

Description:

     Randomised trial of two treatment regimens for lung cancer.  This
     is a standard survival analysis data set.

Usage:

     veteran
     
Format:

       trt:       1=standard 2=test                            
       celltype:  1=squamous,  2=smallcell,  3=adeno,  4=large 
       time:      survival time                                
       status:    censoring status                             
       karno:     Karnofsky performance score (100=good)       
       diagtime:  months from diagnosis to randomisation       
       age:       in years                                     
       prior:     prior therapy 0=no, 10=yes      
```


# Review

##

Nonparametric:

- simple
- does not depend on model fit

. . .

Parametric:

- makes more sense: real hazard rates are not piecewise constant
- better power, *if* the model fits
- can simulate



## The Kaplan-Meier curve, by hand:

```{r byhand}
f1 <- survfit(Surv(time, status) ~ 1, data = lung)

y <- data.frame(time=sort(unique(lung$time[lung$status == 2])))
y$k <- sapply(y$time, function (t) sum(lung$status == 2 & lung$time == t))
y$n <- sapply(y$time, function (t) sum(lung$time > t | (lung$time == t & lung$status == 2)))
y$p <- cumprod(1 - y$k/y$n)

plot(f1)
points(p ~ time, data=y, col='red', pch=20, cex=0.5)
```
