---
output: html_document
---

# Homework, week 12: Population cycling

**Assignment:** Your task is to use Rmarkdown to write a short report, readable by a technically literate person.
The code you used should not be visible in the final report (unless you have a good reason to show it).

**Due:** Submit your work via Canvas by the end of the day (midnight) on Thursday, January 21th.
Please submit **both the Rmd file and the resulting html or pdf file.**
You can work with other members of class, but I expect each of you to write the report yourselves.


## The Problem

In the file [mice_and_foxes.tsv](mice_and_foxes.tsv) you will find (fake) data
consisting of the total number of mice and foxes on a certain large island across 100 years.
We'd like to fit a *Lotka-Volterra*-type model to these data,
in which mice reproduce, but are eaten by foxes,
and fox reproduction rate depends on how many mice they eat.
Skipping over a lot of details,
we'd like to fit the following model of how next year's numbers of mice ($M_{t+1}$) and foxes ($F_{t+1}$)
depend on the current year's numbers ($M_t$ and $F_t$):
$$\begin{aligned}
    r_t &= \exp(- \epsilon F_t / 1000) \\
    M_{t+1} &\sim \Poisson( (\lambda + r_t) M_t ) \\
    F_{t+1} &\sim \Poisson( p F_t + \gamma (1 - r_t) M_t ) .
\end{aligned}$$
In this model, the parameters are:

- $r = \exp(-\epsilon F)$ is the chance that a mouse escapes all foxes when there are $F$ foxes, and so
- $\epsilon$ is the per fox encounter rate, scaled
- $\lambda$ is the per capita mouse fecundity
- $p$ is the probability of survival until the next year for each fox
- $\gamma$ is the conversion rate from mice eaten to baby foxes

Below is some R code that builds, evaluates, and fits this model,
in particular:

1. Simulates from this model
2. Writes a Stan model
3. Evaluates its performance estimating $\lambda$ and $\gamma$
4. Fits the model to real data
5. Simulates dynamics using a few draws from the posterior
6. Looks at the posterior distribution on the parameters.

Your goal is to build on this code to create an Rmarkdown report:
the report should, as usual, explain what is happening in the code
(i.e., explain the method of inference, model validation results, the results in real terms,
and associated uncertainty),
but not show the actual code.
You are also welcome to add code of your own, of course.

*Note:* [This Stan case study](https://mc-stan.org/users/documentation/case-studies/lotka-volterra-predator-prey.html)
addresses a similar question but using very different methods
(that are more confusing, in my opinion, since it fits a differential equation model).

The [Rmd source](HW12_W21.Rmd) to this document is also available.


```{r no_eval, echo=FALSE}
# REMOVE THIS CHUNK TO GET THE CODE TO RUN
knitr::opts_chunk$set(eval=FALSE)
```


```{r sim_code}
# simulation code
generation <- function (mf, p) {
    oM = mf[1]
    oF = mf[2]
    er <- exp( - p$erate * oF / 1000)
    nM <- rpois(1, (p$lambda + er) * oM)
    nF <- rpois(1, p$lsurv * oF + p$conv * (1 - er) * oM)
    return(c(nM, nF))
}

run_sim <- function (mf0, params) {
    MF <- matrix(0, nrow=params$T, ncol=2)
    MF[1,] <- as.vector(mf0)
    for (k in 2:nrow(MF)) {
        MF[k,] <- generation(MF[k-1,], params)
    }
    MF <- data.frame(MF)
    names(MF) <- c("mice", "foxes")
    MF$year = 1750 + 1:params$T
    MF <- MF[,c(3,1,2)]
    return(MF)
}

plot_MF <- function (MF, layout=TRUE, ...) {
    if (layout) layout(t(1:2))
    matplot(MF$year, MF[,2:3], type='l', xlab='time', ylab='population', lty=2, ...)
    legend("topright", lty=1, col=1:2, legend=c("mice", "foxes"))
    plot(foxes ~ mice, data=MF, type='l', xlab='mice', ylab='foxes', ...)
}

# example simulation
set.seed(123)
sim_params <- list(
                   lambda = 0.75, # prey fecundity per year
                   lsurv = 0.4,   # predator survival prob
                   erate = 1,     # pred-prey encounter rate
                   conv = 0.8,    # prey->pred conversion rate
                   T = 100  # number of time steps
               )

plot_MF(run_sim(c(1200, 1000), sim_params), main='simulated data')

# but beware: mice die out with no foxes
bad_params <- sim_params
bad_params$lambda <- 0.2
plot_MF(run_sim(c(1200, 1000), bad_params), main='oops!')
```



```{r setup}
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

MF <- read.table("mice_and_foxes.tsv", header=TRUE)
```

```{r plot_mf}
plot_MF(MF)
```


```{r thestan, cache=TRUE}
pp_block <- "
data {
    int N;
    int M[N];
    int F[N];
}
parameters {
    real<lower=0> lambda;
    real<lower=0, upper=1> lsurv;
    real<lower=0> erate;
    real<lower=0> conv;
}
model {
    vector[N-1] h;
    vector[N-1] l;
    vector[N-1] er;
    h = to_vector(M[1:(N-1)]);
    l = to_vector(F[1:(N-1)]);
    er = exp(-erate * l / 1000);
    M[2:N] ~ poisson((lambda + er) .* h);
    F[2:N] ~ poisson(lsurv * l + conv * (1-er) .* h);
    lambda ~ normal(0, 1);
    erate ~ normal(0, 1);
    conv ~ normal(0, 1);
}
"
pp_model <- stan_model(model_code=pp_block)
```

```{r check_lambda_inference, cache=TRUE, dependson="thestan"}
lambda_vals <- seq(0.5, 0.9, length.out=19)
lambda_results <- data.frame(lambda=lambda_vals)
summary_names <- c("2.5%", "25%", "50%", "75%", "97.5%", "n_eff")
for (x in summary_names) { lambda_results[[x]] <- NA }

for (k in seq_along(lambda_vals)) {
    p <- sim_params
    p$lambda <- lambda_vals[k]
    sim_MF <- run_sim(c(1200, 1000), p)
    # plot_MF(sim_MF)  # uncomment to get visual check on simulations
    sim_pp_fit <- sampling(pp_model,
                       data=list(N=nrow(sim_MF),
                                 M=sim_MF$mice,
                                 F=sim_MF$foxes),
                       iter=1000, chains=3,
                       control=list(max_treedepth=12))
    post_summary <- rstan::summary(sim_pp_fit)$summary
    for (x in summary_names) {
        lambda_results[k, x] <- post_summary["lambda", x]
    }
}

plot(lambda_results$lambda, lambda_results[["50%"]], ylim=range(lambda_results[,1:6]), type='n',
     xlab="true value of lambda", ylab="inferred value")
segments(x0=lambda_results$lambda,
         y0=lambda_results[["2.5%"]],
         y1=lambda_results[["97.5%"]])
segments(x0=lambda_results$lambda,
         y0=lambda_results[["25%"]],
         y1=lambda_results[["75%"]], lwd=2, col='red')
points(lambda_results$lambda, lambda_results[["50%"]], pch=20, cex=2)
abline(0,1)
legend("topleft", pch=c(20, NA, NA), lty=c(NA, 1, 1), lwd=c(NA, 1, 2), col=c("black", "black", "red"),
       legend=c("posterior median", "posterior 50%", "posterior 95%"))
```

```{r check_gamma_inference, cache=TRUE, dependson="thestan"}
gamma_vals <- seq(0.3, 0.5, length.out=19)
gamma_results <- data.frame(gamma=gamma_vals)
for (x in summary_names) { gamma_results[[x]] <- NA }

for (k in seq_along(gamma_vals)) {
    p <- sim_params
    p$conv <- gamma_vals[k]
    sim_MF <- run_sim(c(1200, 1000), p)
    # plot_MF(sim_MF)  # uncomment to get visual check on simulations
    sim_pp_fit <- sampling(pp_model,
                       data=list(N=nrow(sim_MF),
                                 M=sim_MF$mice,
                                 F=sim_MF$foxes),
                       iter=1000, chains=3,
                       control=list(max_treedepth=12))
    post_summary <- rstan::summary(sim_pp_fit)$summary
    for (x in summary_names) {
        gamma_results[k, x] <- post_summary["conv", x]
    }
}

plot(gamma_results$gamma, gamma_results[["50%"]], ylim=range(gamma_results[,1:6]), type='n',
     xlab='true value of gamma', ylab='inferred value')
segments(x0=gamma_results$gamma,
         y0=gamma_results[["2.5%"]],
         y1=gamma_results[["97.5%"]])
segments(x0=gamma_results$gamma,
         y0=gamma_results[["25%"]],
         y1=gamma_results[["75%"]], lwd=2, col='red')
points(gamma_results$gamma, gamma_results[["50%"]], pch=20, cex=2)
abline(0,1)
legend("topleft", pch=c(20, NA, NA), lty=c(NA, 1, 1), lwd=c(NA, 1, 2), col=c("black", "black", "red"),
       legend=c("posterior median", "posterior 50%", "posterior 95%"))
```



```{r fitstan, cache=TRUE, dependson="thestan"}
# ok the real data now
pp_fit <- sampling(pp_model,
                   data=list(N=nrow(MF),
                             M=MF$mice,
                             F=MF$foxes),
                   iter=1000, chains=3,
                   control=list(max_treedepth=12))
```



```{r sim_from_posterior}
post <- rstan::extract(pp_fit, permuted=TRUE)
get_post_params  <- function (k) {
    p <- list(
              lambda = post$lambda[k],
              lsurv = post$lsurv[k],
              erate = post$erate[k],
              conv = post$conv[k],
              T = 100
           )
    return(p)
}
layout(matrix(1:12, nrow=3, byrow=TRUE))
plot_MF(MF, main="real data", layout=FALSE)
for (k in 1:5) {
    p <- get_post_params(k)
    plot_MF(run_sim(c(MF[1,2], MF[1,3]), p), main=sprintf("posterior draw %d", k), layout=FALSE)
}
```

```{r posterior}
stan_hist(pp_fit) # TODO: add titles, labels
```
