
# The data

Downloaded daily case counts
from [the CDC](https://data.cdc.gov/Case-Surveillance/United-States-COVID-19-Cases-and-Deaths-by-State-o/9mfq-cb36),
and subset it to only observations on the 1st, 11th, or 21st of each month,
and the 52 states (fifty plus Puerto Rico and DC):
```{r the_data}
library(lubridate)

x <- read.csv("United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv")
x$submission_date <- mdy(x$submission_date)
    
firsts <- subset(x, day(submission_date) == 1 | day(submission_date) == 11 | day(submission_date) == 21)[,c("submission_date", "state", "new_case")]
names(firsts)[1] <- "date"
by_date <- reshape(firsts, direction = "wide",
        idvar = "state", timevar = "date"
)
names(by_date) <- gsub("new_case.", "", names(by_date))
by_date <- subset(by_date, state %in%
    c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", 
      "FL", "GA", "HI", "IA", "ID", "IL", "IN", "KS", "KY", "LA", 
      "MA", "MD", "ME", "MI", "MN", "MO", "MS", "MT", "NC", "ND", 
      "NE", "NH", "NJ", "NM", "NV", "NY", "OH", "OK", "OR", "PA", "PR", 
      "RI", "SC", "SD", "TN", "TX", "UT", "VA", "VT", "WA", 
      "WI", "WV", "WY")
)
```

# Predicting the future

Let's see how well we can predict the most recent case counts (21 January)
using previous data:

```{r predict}
jan_lm <- lm(`2022-01-21` ~ . - state, data=by_date)

plot(predict(jan_lm), by_date$`2022-01-21`,
     xlab='predicted case numbers', ylab='observed case numbers', type='n')
text(predict(jan_lm), by_date$`2022-01-21`,
     labels=by_date$state)
legend("topleft", bty='n',
       legend=sprintf("r = %f", cor(predict(jan_lm), by_date$`2022-01-21`)))
abline(0, 1)

```

We can predict the case counts *perfectly*!

# Predicting across states

Now let's see how well the case counts in Oregon
are predicted by other states,
using only the 1st and 15th of each month, this time:

```{r bystate}
firsts <- subset(x, day(submission_date) == 1 | day(submission_date) == 15 )[,c("submission_date", "state", "new_case")]
names(firsts)[1] <- "date"
by_state <- reshape(firsts, direction = "wide",
        timevar = "state", idvar = "date"
)
names(by_state) <- gsub("new_case.", "", names(by_state))
by_state <- by_state[, c("date", 
    c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", 
      "FL", "GA", "HI", "IA", "ID", "IL", "IN", "KS", "KY", "LA", 
      "MA", "MD", "ME", "MI", "MN", "MO", "MS", "MT", "NC", "ND", 
      "NE", "NH", "NJ", "NM", "NV", "NY", "OH", "OK", "OR", "PA", "PR", 
      "RI", "SC", "SD", "TN", "TX", "UT", "VA", "VT", "WA", 
      "WI", "WV", "WY")
)]
by_state <- by_state[order(by_state$date),]

OR_lm <- lm(OR ~ . - date, data=by_state)

plot(OR ~ date, data=by_state, type='l', lwd=2,
     ylab="Oregon new cases")
lines(by_state$date, predict(OR_lm), col='red', lty=2, lwd=2)
legend("topleft", lty=c(1, 2), col=c("black", "red"), lwd=c(2,2), legend=c("observed", "predicted using other states"))


```
