---
title: "Permutation test"
author: "Peter Ralph"
date: "1/13/2022"
output: html_document
---



Your turn:

Use a permutation test to assess whether the relation between hair and eye color differs by sex.

To do this, we’ll:

1.    Compute the sum of squared differences between each of the 16 proportions between males and females (4 eye x 4 hair colors). (test statistic)

2.    Compute the same thing, after shuffling up the sex column, a bunch of times.

    Compare the observed value to the distribution under permutations from (2).

We want to write
```
Don't Repeat Yourself
```
code, not
```
Write Everything Twice
```
code!

```{r the_data}
data(HairEyeColor)
haireye <- as.data.frame(HairEyeColor)
names(haireye) <- tolower(names(haireye))
names(haireye)[names(haireye) == "freq"] <- "number"
long_haireye <- haireye[rep(1:nrow(haireye), haireye$number), 
                        c("hair", "eye", "sex")]
stopifnot(nrow(long_haireye) == sum(haireye$number))
````

# Compute the test statistic on the data

```{r test_stat}
tab <- table(long_haireye)
stopifnot(all(tab == HairEyeColor))
# male proportions
pm <- tab[,,"Male"] / sum(tab[,,"Male"])
stopifnot(sum(pm) == 1)
pf <- tab[,,"Female"] / sum(tab[,,"Female"])
stopifnot(sum(pf) == 1)

# consistency check:
for (e in levels(long_haireye$eye)) {
    for (h in levels(long_haireye$hair)) {
        for (g in levels(long_haireye$sex)) {
            p <- with(long_haireye,
                      sum(eye == e & hair == h & sex == g)
                      /
                      sum(sex == g)
            )
            if (g == "Male") {
                stopifnot(pm[h, e] == p)
            } else {
                stopifnot(pf[h, e] == p)
            }
        }
    }
}

summary_stat <- function (tab) {
    pm <- tab[,,"Male"] / sum(tab[,,"Male"])
    stopifnot(sum(pm) == 1)
    pf <- tab[,,"Female"] / sum(tab[,,"Female"])
    stopifnot(sum(pf) == 1)
    ssd <- sum( (pm - pf)^2 )
    return(ssd)
}

stopifnot( sum( (pm - pf)^2 ) == summary_stat(tab) )
```

# Do the same thing after shuffling up the `sex` column a bunch

```{r do_perms}
nperms <- 10000

shuffle <- function(x) {
    x$sex <- sample(x$sex)
    return(x)
}

reps <- replicate(nperms, { 
    newtab <- shuffle(long_haireye)
    summary_stat( table(newtab) )
    }
)
```

# 3. Compare observed value to distribution under permutations

```{r compare}
observed <- summary_stat(table(long_haireye))
hist(reps, breaks=40)
abline(v=observed, col='red', lwd=3)
```

The $p$-value from the permutation test is (approximately)
`r mean(reps > observed)`.
