# Homework 3: Interpret the code

**Assignment:** 
You should use R for your work, and 
submit your work as **both** a final .Rmd file and rendered pdf or html files.

**Due:** Submit your work via Canvas by the end of the day (midnight) on Thursday, January 31. 
You can work with other members of class, but I expect each of you to construct and run all of the scripts
yourself.

```{r data, include=FALSE}
set.seed(23)
ntreat <- 4
ngeno <- 5
truth <- list(mu=200,
              geno=rnorm(ngeno, sd=5), 
              treat=rnorm(ntreat, sd=5), 
              sigma=2.0)
n <- 2000
leafs <- data.frame(geno = letters[sample.int(ngeno, size=n, replace=TRUE)],
                    treat = sample.int(ntreat, size=n, replace=TRUE))
leafs$y <- pmax(0, truth$mu + truth$geno[leafs$geno] + truth$treat[leafs$treat]
                + rnorm(n, sd=truth$sigma))

```

```{r setup}
library(matrixStats)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

plot_cis <- function (Z, ...) {
    # helper function to plot credible intervals, like stan_plot( )
    q <- colQuantiles(Z, probs=c(0.025, 0.25, 0.5, 0.75, 0.975))
    opar <- par(mar=c(par("mar"), 10)[c(1,5,3,4)])
    plot(q[,3], 1:ncol(Z), type='n', 
         xlab='value', ylab='', yaxt='n', xlim=range(q),
         ...)
    segments(x0=q[,1], x1=q[,5], y0=1:ncol(Z))
    segments(x0=q[,2], x1=q[,4], y0=1:ncol(Z), lwd=3, col='red')
    points(q[,3], 1:ncol(Z), pch=20, cex=2)
    axis(2, at=1:ncol(Z), labels=colnames(Z), las=2)
    par(opar)
}
```

## The problem

Here is a dataset ([`leaf_mass.tsv`](leaf_mass.tsv)).

```{r read_data, cache=TRUE}
leafs <- read.table("leaf_mass.tsv", 
                    header=TRUE)
# how many observations?
table(leafs$geno, leafs$treat)
```

```{r raw_data}
# here's the data
boxplot(y ~ geno + treat, 
        main='dry leaf mass', 
        xlab='genotype.treatment',
        col=1:ngeno,
        las=3,
        data=leafs)
abline(v=ngeno * (1:(ntreat-1)) + 0.5, lty=3)
text(x=ngeno * (1:(ntreat)) - 2, y=220, labels=paste("treatment", 1:ntreat))
```

```{r stan_model, cache=TRUE, dependson="read_data"}
anova_model <- stan_model(model_code=" 
data {
    int N;
    int ngeno;
    int ntreat;
    vector[N] leaf;
    int geno[N];
    int treat[N];
}
parameters {
    real mu;
    vector[ngeno] alpha;
    vector[ntreat] beta;
    matrix[ngeno, ntreat] gamma;
    matrix<lower=0>[ngeno, ntreat] sigma;
    real<lower=0> sigma_gamma;
}
model {
    vector[N] muvec;
    vector[N] sigmavec;
    muvec = mu + alpha[geno] + beta[treat];
    for (k in 1:N) {
        muvec[k] += sigma_gamma * gamma[geno[k], treat[k]];
        sigmavec[k] = sigma[geno[k], treat[k]];
    }
    leaf ~ normal(muvec, sigmavec);
    mu ~ normal(205, 10);
    alpha ~ normal(0, 10);
    beta ~ normal(0, 10);
    for (k in 1:ngeno) {
        // using Matt's trick here
        gamma[k] ~ normal(0, 1);
        sigma[k] ~ gamma(20, 10);
    }
    sigma_gamma ~ gamma(0.5, 0.5);
}
")
```

```{r do_sampling, cache=TRUE, dependson="stan_model"}
fit <- sampling(anova_model,
                iter = 1e4,
                data=list(N=nrow(leafs),
                          ngeno=ngeno,
                          ntreat=ntreat,
                          leaf=leafs$y,
                          geno=as.numeric(leafs$geno),
                          treat=as.numeric(leafs$treat)))
print(fit)
```

```{r FIXME}
leafs$treat <- factor(leafs$treat)
```

```{r get_posteriors}
# now we get the posterior means for each group
samples <- extract(fit)
post_means <- as.vector(samples$mu) + samples$gamma
for (k in 1:ngeno) {
    post_means[,k,] <- post_means[,k,] + samples$alpha[,k]
}
for (k in 1:ntreat) {
    post_means[,,k] <- post_means[,,k] + samples$beta[,k]
}

dim(post_means) <- c(dim(post_means)[1], prod(dim(post_means)[2:3]))
colnames(post_means) <- outer(levels(leafs$geno), levels(leafs$treat), paste, sep=".")
post_means <- as.data.frame(post_means)
post_quantiles <- colQuantiles(as.matrix(post_means), probs=c(0.025, 0.25, 0.5, 0.75, 0.975))

# here we plot the data with the posterior estimates of the group mean on top
yfact <- factor(paste(leafs$geno, leafs$treat, sep="."), 
                levels=colnames(post_means))
plot(y ~ as.numeric(yfact), 
     main='dry leaf mass', 
     xlab='genotype.treatment', xaxt='n',
     pch=20, 
     col=adjustcolor(as.numeric(geno), 0.5),
     data=leafs, las=3)
axis(1, at=1:(ngeno * ntreat), labels=levels(yfact), las=3)
segments(x0=1:(ngeno * ntreat), col='red',
         y0=post_quantiles[,1],
         y1=post_quantiles[,5])
segments(x0=1:(ngeno * ntreat), lwd=3,
         y0=post_quantiles[,2],
         y1=post_quantiles[,4])
points(1:(ngeno * ntreat),
       post_quantiles[,3],
       pch=20, cex=2, col='black')
abline(v=ngeno * (1:(ntreat-1)) + 0.5, lty=3)
text(x=ngeno * (1:(ntreat)) - 2, y=220, labels=paste("treatment", 1:ntreat))
```

```{r marginal effects}
post_geno <- samples$alpha
colnames(post_geno) <- paste("alpha", levels(leafs$geno), sep="_")
for (k in 1:ngeno) {
    post_geno[,k] <- post_geno[,k] + samples$mu + rowMeans(samples$gamma[,k,])
}
post_treat <- samples$beta
colnames(post_treat) <- paste("beta", levels(leafs$treat), sep="_")
for (k in 1:ntreat) {
    post_treat[,k] <- post_treat[,k] + samples$mu + rowMeans(samples$gamma[,,k])
}
plot_cis(cbind(post_geno, post_treat))
```
